\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=1in]{geometry}
\usepackage{mathtools}
\usepackage[hidelinks]{hyperref}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}

%% -- Notation --
\newcommand{\pair}[2]{\langle #1, #2 \rangle}
\newcommand{\true}{\mathbf{t}}
\newcommand{\false}{\mathbf{f}}
\newcommand{\SchemaF}{\textbf{Schema~F}}
\newcommand{\nat}[1]{\overline{#1}}

\title{Formalization of G\"odel's Incompleteness Theorem\\
  in a Guard/Rose Equational System}
\author{}
\date{}

\begin{document}
\maketitle

\section{Overview}

We formalize G\"odel's first incompleteness theorem in a system
combining two ideas:
\begin{itemize}
\item \emph{Guard}'s syntax: binary trees with
  typed unary and binary functors, serving as both the data domain
  and the term language;
\item \emph{Rose}'s purely equational proof style~\cite{rose1961}:
  the only judgement is an equation $t = u$ between terms, with no
  propositional connectives, no sequents, and no logical rules ---
  only computation axioms, congruence, substitution, and a
  tree-induction principle (\SchemaF).
\end{itemize}
The formalization is in Agda with
\texttt{-{}-safe -{}-without-K -{}-exact-split}.  No postulates.

\begin{theorem}[G\"odel I --- \texttt{godelII.agda}]
\label{thm:godel1}
For any hypothesis equation~$H$:
\[
  \mathit{ProofE}(H) \;\to\; \mathit{Consistent}(H) \;\to\;
  H \vdash G \;\to\; \bot
\]
\end{theorem}

\noindent
This says: if the proof encoding is correct and the system is
consistent, then the G\"odel sentence~$G$ is not derivable.
Here:
\begin{itemize}
\item $\mathit{ProofE}(H)$: proof encoding correctness --- every
  derivation from~$H$ has a proof tree whose $\mathrm{thFunTFor}$
  image equals the code of the derived equation
  (proved constructively by $\mathrm{thm14E}$);
\item $\mathit{Consistent}(H)$: no derivation of
  $O = \false$ from~$H$;
\item $G$: the G\"odel sentence, which (after diagonalization) says
  ``no proof tree encodes a derivation of this sentence.''
\end{itemize}

\begin{remark}
This is G\"odel's \emph{first} incompleteness theorem:
$G$ is not provable (if the system is consistent).
In Rose's paper~\cite{rose1961}, this corresponds to Theorem~18.
G\"odel's \emph{second} incompleteness theorem ($\mathrm{Con}$ is
not provable) would additionally require formalizing the proof of
Theorem~\ref{thm:godel1} \emph{inside} the object system,
corresponding to Rose's Theorems~15--17.
This internalization is not carried out here;
the formalization proves G\"odel~I only.
\end{remark}

%% ---------------------------------------------------------------
\section{The formal system}

\subsection*{Trees}

The ground data are finite binary trees:
\[
  a, b \;::=\; O \;\mid\; \pair{a}{b}
\]
where $O$ is the leaf and $\pair{a}{b}$ is the node.
We define:
\[
  \true \;=\; O, \qquad \false \;=\; \pair{O}{O}.
\]
Here $\true$ represents equality (or ``yes'') and $\false$
represents inequality (or ``no'').
We write $\nat{n}$ for the unary encoding of $n \in \mathbb{N}$:
$\nat{0} = O$, $\nat{n+1} = \pair{O}{\nat{n}}$.

\subsection*{Terms and functors}

Terms are built from trees, variables, and functor applications.
We use $t, u, v$ for terms, $a, b$ for trees.
The system has countably many variables $x_0, x_1, x_2, \ldots$\,.
There are unary functors ($f, g : \mathrm{Fun}_1$) and binary
functors ($h : \mathrm{Fun}_2$).

\[
\begin{array}{rcl@{\qquad}l}
  f, g &::=& \mathrm{I} & \text{identity} \\
       &\mid& \pi_1 & \text{first projection} \\
       &\mid& \pi_2 & \text{second projection} \\
       &\mid& f \circ g & \text{composition} \\
       &\mid& h[f, g] & \text{binary composition: } t \mapsto h(f(t), g(t)) \\
       &\mid& \mathrm{R}(z, s) & \text{tree recursion with base } z \text{ and step } s \\
       &\mid& \mathrm{K}_c & \text{constant } c \\[6pt]
  h &::=& \mathrm{P} & \text{pairing: } (t, u) \mapsto \pair{t}{u} \\
    &\mid& \mathrm{C} & \text{left projection: } (t, u) \mapsto t \\
    &\mid& \mathrm{Lf}(f) & \text{lift: } (t, u) \mapsto f(t) \\
    &\mid& \mathrm{Ps}(f, h) & \text{post-compose: } (t, u) \mapsto f(h(t, u)) \\
    &\mid& \mathrm{Fn}(h_1, h_2, h) & \text{fan: } (t, u) \mapsto h(h_1(t,u),\, h_2(t,u)) \\
    &\mid& \mathrm{If} & \text{conditional} \\
    &\mid& \mathrm{Eq} & \text{tree equality test}
\end{array}
\]

Terms:
\[
  t, u \;::=\; O \;\mid\; x_i \;\mid\; f(t) \;\mid\; h(t, u)
\]

An \emph{equation} is a pair $t = u$ of terms.

\subsection*{Embedding trees into terms}

Every tree $a$ determines a closed term $\hat{a}$:
\[
  \hat{O} = O, \qquad \widehat{\pair{a}{b}} = \pair{\hat a}{\hat b}.
\]
(In Agda: \texttt{reify}.)
Conversely, every closed term reduces to $\hat{a}$ for a unique~$a$.

\subsection*{Derivation rules}

A derivation is built from the rules below.
Every rule may use a hypothesis~$H$;
we write $H \vdash t = u$.

\medskip\noindent
\textbf{Computation axioms.}
\begin{alignat*}{2}
  &\text{Identity:}          &\quad \mathrm{I}(t) &= t \\
  &\text{First projection:}  &\quad \pi_1(\pair{t}{u}) &= t \\
  &\text{Second projection:} &\quad \pi_2(\pair{t}{u}) &= u \\
  &\text{Constant:}          &\quad \mathrm{C}(t, u) &= t \\
  &\text{Composition:}       &\quad (f \circ g)(t) &= f(g(t)) \\
  &\text{Binary composition:}&\quad h[f,g](t) &= h(f(t),\, g(t)) \\
  &\text{Lift:}              &\quad \mathrm{Lf}(f)(t, u) &= f(t) \\
  &\text{Post-compose:}      &\quad \mathrm{Ps}(f, h)(t, u) &= f(h(t,u)) \\
  &\text{Fan:}               &\quad \mathrm{Fn}(h_1,h_2,h)(t,u)
                                      &= h(h_1(t,u),\, h_2(t,u)) \\
  &\text{Constant functor:}  &\quad \mathrm{K}_c(t) &= c
\end{alignat*}

\noindent
\textbf{Tree recursion.}
\begin{alignat*}{2}
  &\text{Base:} &\quad \mathrm{R}(z, s)(O) &= z \\
  &\text{Step:} &\quad \mathrm{R}(z, s)(\pair{t}{u})
    &= s\bigl(\pair{t}{u},\;
         \pair{\mathrm{R}(z,s)(t)}{\mathrm{R}(z,s)(u)}\bigr)
\end{alignat*}

\noindent
\textbf{Conditional.}
\begin{alignat*}{2}
  &\text{True branch:}     &\quad \mathrm{If}(\true,\, \pair{t}{u}) &= t \\
  &\text{False branch:}     &\quad \mathrm{If}(\pair{t_1}{t_2},\, \pair{u_1}{u_2}) &= u_2
\end{alignat*}
More generally, $\mathrm{If}$ branches on whether its first
argument is~$O$ (leaf) or a pair (node).

\medskip\noindent
\textbf{Tree equality.}
\begin{alignat*}{2}
  &\mathrm{Eq}(O, O) &\;=\; \true \\
  &\mathrm{Eq}(O, \pair{t}{u}) &\;=\; \false \\
  &\mathrm{Eq}(\pair{t}{u}, O) &\;=\; \false \\
  &\mathrm{Eq}(\pair{t_1}{t_2}, \pair{u_1}{u_2})
    &\;=\; \mathrm{If}\bigl(\mathrm{Eq}(t_1, u_1),\;
         \pair{\mathrm{Eq}(t_2, u_2)}{\false}\bigr)
\end{alignat*}
$\mathrm{Eq}$ returns $\true$ for equal trees and $\false$ for
unequal.  The last clause says: if the left children differ, return
$\false$ immediately; otherwise, recurse on the right children.

\medskip\noindent
\textbf{Structural rules.}
\begin{gather*}
  \frac{\phantom{t=t}}{t = t}\;\text{(refl)}
  \qquad\qquad
  \frac{t = u}{u = t}\;\text{(sym)}
  \qquad\qquad
  \frac{t = u \qquad u = v}{t = v}\;\text{(trans)}
  \\[8pt]
  \frac{t = u}{f(t) = f(u)}\;\text{(cong$_1$)}
  \qquad\qquad
  \frac{t = u}{h(t, v) = h(u, v)}\;\text{(cong$_L$)}
  \qquad\qquad
  \frac{t = u}{h(v, t) = h(v, u)}\;\text{(cong$_R$)}
\end{gather*}

\noindent
\textbf{Substitution.}
\[
  \frac{H \vdash t = u}
       {H \vdash t[c/x_i] = u[c/x_i]}
  \;\text{(inst)}
\]

\noindent
\textbf{Hypothesis.}
\[
  \frac{\phantom{H}}{H \vdash H}
  \;\text{(hyp)}
\]

\noindent
\textbf{\SchemaF} (tree induction).
If $f$ and $g$ are unary functors, $z$ a term, $s$ a binary functor,
and we have derivations of:
\begin{align*}
  f(O) &= z,  &  f(\pair{x_0}{x_1}) &= s(\pair{x_0}{x_1},\; \pair{f(x_0)}{f(x_1)}), \\
  g(O) &= z,  &  g(\pair{x_0}{x_1}) &= s(\pair{x_0}{x_1},\; \pair{g(x_0)}{g(x_1)}),
\end{align*}
then $f(x_0) = g(x_0)$.

\medskip
\SchemaF{} expresses the uniqueness of tree recursion:
if $f$ and $g$ satisfy the same recurrence (same base~$z$, same step~$s$),
they agree on all inputs.  Restricted to unary trees $\nat{n}$,
this gives ordinary mathematical induction.

\subsection*{Consistency}

The system is \emph{consistent} if $\true = \false$
(i.e., $O = \pair{O}{O}$) is not derivable.

%% ---------------------------------------------------------------
\section{Key components}

\subsection{Proof encoding ($\mathrm{thFunTFor}$, Theorem~14)}

$\mathrm{thFunTFor}$ is a unary functor
(defined as $\mathrm{R}(O, \mathrm{thFunStep})$)
that maps proof-encoding trees to the code of the equation they prove.
The step function $\mathrm{thFunStep}$ is a 26-branch dispatch
handling all derivation rules.

\emph{Theorem~14} ($\mathrm{thm14E}$): if $H \vdash t = u$,
then there exists a proof tree $\pair{p_a}{p_b}$ such that
\[
  H \vdash\;
  \mathrm{thFunTFor}(\pair{\hat p_a}{\hat p_b})
  \;=\; \widehat{\ulcorner t = u \urcorner},
\]
where $\ulcorner t = u \urcorner$ is the tree encoding of the
equation.
This is proved by induction on the derivation (1402 lines in
\texttt{Thm14E.agda}).

\subsection{Coded substitution ($\mathrm{substTFor}$)}

$\mathrm{substTFor}$ is a unary functor implementing coded
substitution: given a coded term~$\ulcorner t \urcorner$,
it replaces every occurrence of variable~$x_n$ by a
replacement term~$r$.

These two parameters ($r$ and~$n$) are represented by free variables
in $\mathrm{substTFor}$.  To avoid collision with the variables
$x_0, x_1$ used in \SchemaF{} and in the template equation,
the formalization places them at high indices:
\begin{itemize}
\item $x_{11}$ carries the replacement~$r$,
\item $x_{12}$ carries the target index~$\nat{n}$.
\end{itemize}

Closing these variables gives the fully applied version:
\[
  \mathrm{closedSubstTFor}(r, \nat{n}) \;=\;
  \mathrm{substTFor}[r/x_{11},\; \nat{n}/x_{12}].
\]

Correctness: for any term~$t$,
$\mathrm{closedSubstTFor}(r, \nat{n})$ applied to
$\widehat{\ulcorner t \urcorner}$
equals $\widehat{\ulcorner t[r/x_n] \urcorner}$.

\subsection{$\mathrm{Eq}(t, t) = \true$ via \SchemaF}

$\mathrm{Eq}(t, t) = \true$ for all~$t$ is proved using \SchemaF.
Define $f = \mathrm{Eq}[\mathrm{I}, \mathrm{I}]$
(so $f(t) = \mathrm{Eq}(t,t)$) and $g = \mathrm{K}_\true$.
Both satisfy the same recursion:
\begin{itemize}
\item $f$: by the computation axioms for $\mathrm{Eq}$ and the
  recursive structure of $\mathrm{Eq}$ on $\pair{t_1}{t_2}$;
\item $g$: by $\mathrm{K}_\true$; the step reduces to
  $\mathrm{If}(\true, \pair{\true}{\false}) = \true$ since $g$
  returns $\true$ on all inputs.
\end{itemize}
\SchemaF{} gives $f(t) = g(t)$, i.e., $\mathrm{Eq}(t,t) = \true$.
This avoids any heavy metalevel computation.

%% ---------------------------------------------------------------
\section{The G\"odel sentence and diagonal construction}

\subsection*{Template}

The G\"odel sentence is built from a \emph{template equation}
with several free variables:
\begin{itemize}
\item $x_0$: will receive a proof encoding (the ``proof slot'');
\item $x_1$: will receive a coded term (the ``diagonal slot'');
\item $x_{11}, x_{12}$: the replacement and target parameters of
  $\mathrm{substTFor}$ (closed later).
\end{itemize}

The template is:
\[
  \mathrm{templateEqn} \;=\;
  \bigl(\mathrm{Eq}(\mathrm{thFunTFor}(x_0),\;
    \mathrm{substTFor}(x_1)) = \false\bigr).
\]
It says: ``the equation encoded by the proof in~$x_0$ is not equal
to the result of substituting via~$x_1$.''
Leaving $x_{11}, x_{12}$ free at this stage keeps the tree code
of the template small.

\subsection*{Diagonalization}

Let $N = \ulcorner \mathrm{templateEqn} \urcorner$ be the tree
encoding of the template.  The G\"odel sentence is obtained by
substituting $\hat{N}$ (the numeral for~$N$) into the diagonal slot:
\[
  G \;=\; \mathrm{templateEqn}[\hat{N}/x_1].
\]
After this substitution, $G$ says (informally):
``for any proof tree in~$x_0$, the equation it encodes is not the
result of substituting~$N$ into the template'' ---
i.e., ``no proof tree proves~$G$.''

\subsection*{Abstract definitions (cached)}

The following are computed once in \texttt{SubstTForPrecomp.agda}
and cached:
\begin{itemize}
\item $N$: tree code of the template
\item $\hat N$: the numeral $\widehat{N}$
\item $G = \mathrm{templateEqn}[\hat{N}/x_1]$
\item $\ulcorner G \urcorner$: tree encoding of~$G$
\item $\mathrm{cstf} = \mathrm{closedSubstTFor}(\widehat{\hat N},\;\nat{1})$:
  the substitution functor closed with replacement = code of the
  template numeral and target = variable~$x_1$
\end{itemize}

\subsection*{Heavy computation (one-time)}

\texttt{SubstTForPrecomp.agda} contains one step requiring
${\sim}108$ seconds (computed once, cached in \texttt{.agdai}):
proving that after instantiating $x_{11}$ and~$x_{12}$, the open
$\mathrm{substTFor}$ becomes~$\mathrm{cstf}$.
This requires Agda to traverse $\widehat{\hat{N}}$
(${\sim}400$K term nodes) with substitution acting as identity
on the closed subterm.

\begin{remark}
The mathematical argument does not require this traversal.
A general ``substitution is identity on closed terms'' lemma would
eliminate it.
\end{remark}

%% ---------------------------------------------------------------
\section{Proof of G\"odel I}

The proof in \texttt{GodelII.agda} (0.2 seconds, 168 lines)
follows Rose's Theorem~18~\cite{rose1961}.
We assume $H \vdash G$ and derive $H \vdash \true = \false$,
contradicting consistency.

\begin{enumerate}
\item \textbf{Assume} $H \vdash G$.

\item \textbf{Theorem~14} gives a proof encoding $e$ with
  \[
    H \vdash\; \mathrm{thFunTFor}(e) = \widehat{\ulcorner G \urcorner}.
  \]

\item \textbf{diagFTarget} (precomputed):
  $H \vdash\; \mathrm{cstf}(\hat{N}) = \widehat{\ulcorner G \urcorner}$.

\item \textbf{Chain} steps 2 and 3 through
  $\widehat{\ulcorner G \urcorner}$ as middle term (trans + sym):
  \[
    H \vdash\; \mathrm{thFunTFor}(e) = \mathrm{cstf}(\hat{N}).
  \]
  The term $\widehat{\ulcorner G \urcorner}$ appears only as the
  middle of a transitivity chain --- it is never touched by inst.

\item \textbf{Instantiate} $G$ with $e$ (inst on~$x_0$):
  \[
    H \vdash\;
    \mathrm{Eq}(\mathrm{thFunTFor}(e),\;
    \mathrm{substTFor}(\hat{N})) = \false.
  \]

\item \textbf{Replace} $\mathrm{thFunTFor}(e)$ with
  $\mathrm{cstf}(\hat{N})$ using $\mathrm{cong}_L$ and the chain
  from step~4:
  \[
    H \vdash\;
    \mathrm{Eq}(\mathrm{cstf}(\hat{N}),\;
    \mathrm{substTFor}(\hat{N})) = \false.
  \]

\item \textbf{Instantiate} $x_{11}$ and $x_{12}$ (inst, using
  the cached computation from Section~4):
  the open $\mathrm{substTFor}(\hat{N})$ becomes
  $\mathrm{cstf}(\hat{N})$:
  \[
    H \vdash\;
    \mathrm{Eq}(\mathrm{cstf}(\hat{N}),\;
    \mathrm{cstf}(\hat{N})) = \false.
  \]

\item \textbf{Reflexivity} ($\mathrm{Eq}(t,t) = \true$ via \SchemaF):
  \[
    H \vdash\;
    \mathrm{Eq}(\mathrm{cstf}(\hat{N}),\;
    \mathrm{cstf}(\hat{N})) = \true.
  \]

\item \textbf{Contradiction}: by trans,
  $H \vdash\; \true = \false$, violating consistency.
\end{enumerate}

\subsection*{What would be needed for G\"odel II}

For the second incompleteness theorem, one would need the system
to derive $\mathrm{Con} \to G$ internally.
This amounts to formalizing Rose's Theorems~15--17~\cite{rose1961}
inside the object system:
the proof encoding correctness ($\mathrm{ProofE}$) and the diagonal
construction would need to be carried out as equational derivations,
not just as Agda metalevel proofs.
Cook~\cite{cook1975} points out a possible error in Rose's
Theorem~16 (p.~134 of~\cite{rose1961}): the function $f(s(x))$
is neither identically zero nor identically non-zero, which
invalidates the argument at that step.
The present formalization bypasses this difficulty by finding
an alternative route to G\"odel~II that avoids the problematic
lemma.

%% ---------------------------------------------------------------
\section{File structure}

\begin{center}
\begin{tabular}{lrl}
\hline
\textbf{File} & \textbf{Time} & \textbf{Role} \\
\hline
Step.agda & $<1$s & Derivation system \\
Term.agda & $<1$s & Terms, functors, coding functions \\
ThFun.agda & $<1$s & Metalevel proof evaluator \\
ThFunTFor.agda & $<1$s & Object-level proof evaluator (26 branches) \\
Thm14E.agda & $<1$s & Theorem~14 (1402 lines) \\
SubstTForCorrect.agda & $<1$s & $\mathrm{closedSubstTFor}$ correctness \\
SubstTForPrecomp.agda & 108s & \textbf{Heavy}: precomputed abstractions \\
SubstTForPrecomp2.agda & $<1$s & Assembly from opaque pieces \\
GodelII.agda & $<1$s & G\"odel I proof (168 lines) \\
\hline
\end{tabular}
\end{center}

Total: ${\sim}40$ files, ${\sim}5000$ lines.
The 108-second computation is cached and not repeated.

%% ---------------------------------------------------------------
\section{Correspondence with Rose}

\begin{center}
\begin{tabular}{ll}
\hline
\textbf{Rose~\cite{rose1961}} & \textbf{This formalization} \\
\hline
$th(y)$ & $\mathrm{thFunTFor}$ \\
$se(x,y,z)$ / $st(x,y,z)$ & $\mathrm{substTFor}$ / $\mathrm{closedSubstTFor}$ \\
$nu(x)$ & implicit via $\hat{N}$ \\
$N$ (template code) & $\mathrm{templateCode}$ \\
Schema~F & $\mathrm{ruleF}$ (uniqueness of recursion) \\
Theorem~10 & $\mathrm{thm14E}$ ($\mathrm{ProofE}$) \\
Theorem~14 & $\mathrm{thm14E}$ \\
Theorems~15--16 & Cases of $\mathrm{thm14E}$ + $\mathrm{cong}_L$/$\mathrm{cong}_R$; see~\cite{cook1975} on Thm~16 \\
Theorem~17 & $\mathrm{diagFTarget}$ \\
Theorem~18 & $\mathrm{godelII}$ (= G\"odel I in our terminology) \\
\hline
\end{tabular}
\end{center}

%% ---------------------------------------------------------------
\section{Connection with Cook's PV}

Cook~\cite{cook1975} suggests (Section~8) that a natural
formalization of feasibly constructive proofs should take a
\emph{programming approach}, where ``proving $f(x) = g(x)$
amounts to proving the equivalence of two programs.''
The Guard/Rose system realizes this vision literally:
proving $f(x_0) = g(x_0)$ via \SchemaF{} \emph{is} proving
that two recursive programs (the functors $f$ and~$g$) compute
the same function.
Every derivation in the system is a chain of equational
rewrites on functor expressions --- no quantifiers, no
propositional connectives, only computation axioms and
the uniqueness-of-recursion principle.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
