\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{xcolor}

\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue},
  commentstyle=\color{gray},
  showstringspaces=false,
  breaklines=true,
}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{proposition}[theorem]{Proposition}

\title{Formalization of G\"odel's Incompleteness Theorems\\
  in Basic Recursive Arithmetic (BRA)}
\author{Summary of an Agda development}
\date{May 2026}

\begin{document}
\maketitle

\begin{abstract}
We present a fully formalized proof of both G\"odel's First and Second
Incompleteness Theorems in a tree-based variant of Guard's Basic
Recursive Arithmetic (BRA), executed in Agda~2.8 with
\verb|--safe --without-K --exact-split|.  The development contains
\textbf{zero postulates and zero holes}; \verb|godelII : Deriv Con -> Deriv bot|
typechecks end-to-end against the abstract Goedel~II skeleton.

There is a single \verb|Term| datatype (no separate \verb|Tree|
datatype), and ``codes'' are exactly the value-shaped Terms certified
by an \verb|IsValue : Term -> Set| predicate (built only from $O$ and
\verb|ap2 Pair|).  The map \verb|reify : Tree -> Term|, written
explicitly throughout the codebase for clarity at code/term
boundaries, is the identity.

The headline results are:
\begin{itemize}
  \item $\mathit{thm11} : \mathit{Deriv}\;G \to \mathit{Deriv}\;\bot$
        (G\"odel~I);
  \item $\mathit{godelII} : \mathit{Deriv}\;\mathit{Con} \to \mathit{Deriv}\;\bot$
        (G\"odel~II).
\end{itemize}
The development follows Guard's 1963 lecture
notes~\cite{guard1963}.  The intermediate construction is a single
first-order combinator $\mathit{constr5} : \mathit{Fun}_1$ realising
Guard's ``$F(x)$'' (page~17); each step is given by an explicit
equational derivation.

A key conceptual point this note emphasises (Section~\ref{sec:internal})
is that intermediate steps of Theorem~14's chain produce
\emph{equational} \verb|Deriv|s of the form
\verb|atomic (eqn (thmT t) u)| where \verb|u| is \emph{not}
\verb|codeFormula P| for any $P : \mathit{Formula}$.  Only the final
\verb|step5| conclusion has \verb|u = codeFormula bot| (with $\bot$
genuinely closed).  Guard's chain operates one level lower than
Hilbert--Bernays--L\"ob: instead of internal provability claims about
specific formulas, it manipulates substituted-codes through standard
BRA inference rules.  This makes the entire proof internal and
equational --- a feature of the framework, not a gap.
\end{abstract}

\section{Introduction}

The aim is a feasible, ASCII-only, syntax-first Agda formalization of
G\"odel's two incompleteness theorems following Guard's 1963 system of
\emph{Basic Recursive Arithmetic}~\cite{guard1963}.

\subsection{Single-layer Term language with $\mathit{IsValue}$}
\label{subsec:layers}

In BRA2 there is one syntactic layer.

\paragraph{The \texttt{Term} language.}
\begin{itemize}
  \item $\mathit{Term}$: built from $O$ (leaf), $\mathit{var}\;n$,
        $\mathit{ap}_1\;f\;t$, $\mathit{ap}_2\;g\;t_1\;t_2$.
  \item Function symbols: $\mathit{Fun}_1$
        ($\{I, \mathit{Fst}, \mathit{Snd}, \mathit{Comp}, \mathit{Comp}_2, Z\}$,
        plus $\mathit{KT}\,\_$ as a defined function), $\mathit{Fun}_2$
        ($\{\mathit{Pair}, \mathit{Const}, \mathit{Lift}, \mathit{Post}, \mathit{Fan}, \mathit{IfLf}, \mathit{TreeEq}, \mathit{treeRec}\}$).
  \item $\mathit{Formula}$: $\mathit{atomic}\;(\mathit{eqn}\;\ell\;r)$, $\mathit{not}\;\_$, $\_\,\mathit{imp}\,\_$.
  \item $\mathit{Deriv}\;F$: inductive family of derivations.  Constructors
        for the algebraic axioms ($\mathit{ax}_*$), the equality axioms
        (Guard's Ax~4--7), three propositional axioms ($\mathit{axK}, \mathit{axS}, \mathit{axNeg}$
        in Lukasiewicz form), modus ponens ($\mathit{mp}$), substitution
        ($\mathit{ruleInst}$), and binary tree induction ($\mathit{ruleIndBT}$).
\end{itemize}

\paragraph{Codes are value-shaped Terms.}
``Codes'' are Terms whose value-shape is certified by
\[
  \mathit{IsValue} : \mathit{Term} \to \mathit{Set},
\]
with constructors
$\mathit{valO} : \mathit{IsValue}\;O$ and
$\mathit{valNd}\;a\;b\;v_a\;v_b : \mathit{IsValue}\;(\mathit{ap}_2\;\mathit{Pair}\;a\;b)$
(when $\mathit{IsValue}\;a$ and $\mathit{IsValue}\;b$).

The encoding maps now have type
\[
  \mathit{code} : \mathit{Term} \to \mathit{Term}, \qquad
  \mathit{codeFormula} : \mathit{Formula} \to \mathit{Term},
\]
together with structural-soundness lemmas
$\mathit{code\_isValue} : (t : \mathit{Term}) \to \mathit{IsValue}\;(\mathit{code}\;t)$,
$\mathit{codeFormula\_isValue}$, $\mathit{codeF}_1\_\mathit{isValue}$,
$\mathit{codeF}_2\_\mathit{isValue}$, and $\mathit{natCode\_isValue}$
(BRA2/Term.agda, BRA2/Formula.agda).  Every primitive constructor of an
encoding produces a value-shaped Term.

\paragraph{Notational aliases \texttt{Tree}, \texttt{lf}, \texttt{nd}, \texttt{reify}.}
For continuity with Guard's prose, BRA2/Term.agda exposes
\[
  \mathit{Tree} \;:=\; \mathit{Term}, \quad
  \mathit{lf} \;:=\; O, \quad
  \mathit{nd}\;a\;b \;:=\; \mathit{ap}_2\;\mathit{Pair}\;a\;b, \quad
  \mathit{reify}\;t \;:=\; t.
\]
The Layer-1/Layer-2 distinction is recorded as the $\mathit{IsValue}$
predicate; syntactically the layers are unified.

\paragraph{Reflection layer (object-level operations on codes).}
Functors that reason about BRA syntax internally:
\begin{itemize}
  \item $\mathit{cor} : \mathit{Fun}_1$: defined as $\mathit{Rec}\;\mathit{stepCor}$
        (BRA2/Cor.agda).  Spec:
        $\mathit{corOfReify} : (t : \mathit{Term}) \to \mathit{IsValue}\;t \to
          \mathit{Deriv}\;(\mathit{ap}_1\;\mathit{cor}\;t = \mathit{code}\;t)$.
        Operationally, $\mathit{cor}$ takes a value-shaped Term (= a
        ``numeral'') and returns its code.
  \item $\mathit{sub} : \mathit{Fun}_2$, $\mathit{subT} : \mathit{Fun}_2$:
        coded substitution (BRA2/Sub.agda, BRA2/SubT.agda), now indexed
        by $\mathit{IsValue}$ on the substituent.
  \item $\mathit{thmT} : \mathit{Fun}_1$: the proof-checker
        (Section~\ref{sec:thmT}).
\end{itemize}

\paragraph{Reflection theorem.}
\[
  \mathit{encode} : (P : \mathit{Formula}) \to \mathit{Deriv}\;P \to
    \Sigma\;\mathit{Term}\;\bigl(\lambda y.\;
      \Sigma\;(\mathit{IsValue}\;y)\;\bigl(\lambda \_.\;\mathit{Deriv}\;
        (\mathit{thmT}\;y = \mathit{codeFormula}\;P)\bigr)\bigr).
\]
The $\mathit{IsValue}\;y$ field certifies that every encoded proof
tree is value-shaped, so $\mathit{cor}\;y$ provably reduces to
$\mathit{code}\;y$ at the witness returned by $\mathit{encode}$.  This
certification is exactly what discharges the parametric \verb|cor x|
in the chain at the closure (see Section~\ref{sec:internal}).

\subsection{Notational convention: code-quotes and hats}
\label{subsec:codequote}

\paragraph{Code-quotes $\ulcorner \cdot \urcorner$.}
We use $\ulcorner E \urcorner$ throughout for ``the code (value-shaped
Term) of $E$'':
\[
  \ulcorner t \urcorner := \mathit{code}\;t \quad (t : \mathit{Term}), \qquad
  \ulcorner F \urcorner := \mathit{codeFormula}\;F \quad (F : \mathit{Formula}).
\]
Whenever a Guard-style expression mixes object-level and encoded
material (e.g.\ $f(\underline{x})$), we keep the code-quote explicit:
the formal object is the Term $\ulcorner f(\underline{x}) =
\underline{f(x)}\urcorner$, never the unquoted equation.  See
Section~\ref{subsec:underline-caveat} for the full caveat.

\paragraph{Hats $\hat\,$ and $\widehat{\,}$.}
We write $\hat x$ as a shorthand for $\mathit{cor}\;x$, the
numeral-encoding combinator applied to a Term $x$.  This lets us write
e.g.\ $\mathit{th}\;\hat x$ for $\mathit{ap}_1\;\mathit{thmT}\;(\mathit{ap}_1\;\mathit{cor}\;x)$
and $\mathit{sub}\;\hat x\;\hat y$ for $\mathit{ap}_2\;\mathit{sub}\;
(\mathit{ap}_1\;\mathit{cor}\;x)\;(\mathit{ap}_1\;\mathit{cor}\;y)$.

For a Guard-style expression $E$, $\widehat{E}$ denotes the
\emph{substituted-code} of $E$: the Term obtained by taking the code
of $E$'s pattern (with variable slots in place of $\hat{}$-marked
positions) and $\mathit{sb}$-substituting the $\hat\,$-encoded numerals
into those slots.  E.g.\ $\widehat{\mathit{th}\;\hat x} =
\mathit{ap}_2\;\mathit{Pair}\;\mathit{tagAp}_1\;
(\mathit{ap}_2\;\mathit{Pair}\;(\mathit{codeF}_1\;\mathit{thmT})\;
(\mathit{ap}_1\;\mathit{cor}\;x))$.  At a free Term variable $x$, this
substituted-code is in general not $\ulcorner P\urcorner$ for any
$P : \mathit{Formula}$ (Section~\ref{sec:internal} discusses this in
detail); for value-shaped $x$ it BRA-reduces to a genuine
codeFormula via $\mathit{corOfReify}$.

\section{The provability predicate $\mathit{thmT}$}
\label{sec:thmT}

The G\"odel-style ``$\mathit{th}$'' provability predicate is realized
in BRA by a single $\mathit{Fun}_1$-functor:
\[
  \mathit{thmT} \;=\; \mathit{Rec}\;\mathit{stepProtoWrapped} \;:\; \mathit{Fun}_1.
\]
Here $\mathit{stepProtoWrapped}$ is a $\mathit{Fun}_2$ assembled from a
40-deep linear $\mathit{IfLf}$ cascade (BRA2/Thm/StepProto.agda and
BRA2/Thm/ThmT.agda), one branch per primitive proof rule: ten
equational axioms of Group~I, eight Group~II axioms, nine
equality and propositional rules, and three rules of inference
($\mathit{mp}$, $\mathit{ruleInst}$, $\mathit{ruleIndBT}$).

The \emph{key} property of $\mathit{thmT}$ is the encoding-soundness
lemma $\mathit{encode}$ above, proved by induction on the
$\mathit{Deriv}$ derivation in BRA2/Thm/ThmT.agda's $\mathit{encodeRich}$.

\subsection{Self-substitution closure}

$\mathit{thmT}$ is a closed combinator:
\[
  \mathit{thClosed} : \forall x\;r,\quad \mathit{substF}_1\;x\;r\;\mathit{thmT} = \mathit{thmT}
\]
and its encoding is var-free:
\[
  \mathit{codeF1Th\_noVar} : \forall x\;\mathit{repl},\quad
    \mathit{codedSubst}\;\mathit{repl}\;(\mathit{code}\;(\mathit{var}\;x))\;(\mathit{codeF}_1\;\mathit{thmT})
    = \mathit{codeF}_1\;\mathit{thmT}.
\]

\section{G\"odel's First Incompleteness Theorem (Theorem~11)}

\subsection{The diagonal sentence}

Following Guard, define
\begin{align*}
  F_{\mathrm{pre}} \;&:=\; \neg \big(\mathit{thmT}\;v_1 \;=\; \mathit{sub}\;v_0\;v_0\big),\\
  j \;&:=\; \mathit{codeFormula}\;F_{\mathrm{pre}} \quad \text{(a closed value-shaped Term)},\\
  G \;&:=\; F_{\mathrm{pre}}[v_0 := j].
\end{align*}
The sentence $G$ has $v_1$ free; informally, $G$ asserts of itself
that no proof tree $v_1$ proves the formula whose code is
$j[v_0 := j]$.

The certificate $j\_\mathit{isValue} : \mathit{IsValue}\;j$ is
exposed inside the \verb|Diagonal| abstract block
(BRA2/Thm11Diagonal.agda).  It propagates through
$\mathit{Outer.GodelII}$ and is what makes
\verb|corOfReify j j_isValue| usable downstream.

(For continuity with prior notation, the Agda codebase additionally
defines \verb|i := reify j|; since \verb|reify| is the identity,
$i = j$.  We use $j$ throughout this paper.)

\subsection{Theorem 11}

\begin{theorem}[\texttt{thm11}, G\"odel I]
\label{thm:godel-I}
\(
  \mathit{thm11} : \mathit{Deriv}\;G \to \mathit{Deriv}\;\bot,
\)
where $\bot := \mathit{atomic}\;(\mathit{eqn}\;\mathit{trueT}\;\mathit{falseT})$,
$\mathit{trueT} := O$, $\mathit{falseT} := \mathit{ap}_2\;\mathit{Pair}\;O\;O$.
\end{theorem}

\begin{proof}[Sketch]
From any $\pi : \mathit{Deriv}\;G$ extract its encoding
$(y_d, v_d, E_1)$ where $v_d : \mathit{IsValue}\;y_d$ and
$E_1 : \mathit{Deriv}\;(\mathit{thmT}\;y_d = \mathit{codeFormula}\;G)$.
A separate diagonal lemma (BRA2/Thm11Diagonal.agda) yields
$N : \mathit{Deriv}\;\neg(\mathit{thmT}\;y_d = \mathit{codeFormula}\;G)$,
and $\mathit{ax}_{\mathit{ExFalso}}$ + two $\mathit{mp}$'s collapse $E_1$
and $N$ to $\mathit{Deriv}\;\bot$.
\end{proof}

\section{G\"odel's Second Incompleteness Theorem (Theorem~14)}

\begin{theorem}[\texttt{godelII}, G\"odel II]
\label{thm:godel-II}
\(
  \mathit{godelII} : \mathit{Deriv}\;\mathit{Con} \to \mathit{Deriv}\;\bot,
\)
where $\mathit{Con} := \neg(\mathit{thmT}\;v_0 = \mathit{codeFormula}\;\bot)$.
\end{theorem}

\subsection{Caveat: ``$f(\underline{x}) = \underline{f(x)}$'' is not a BRA formula}
\label{subsec:underline-caveat}

A structurally crucial point, easy to miss in Guard's notation: the
expression
\[
  f(\underline{x}) \;=\; \underline{f(x)}
\]
is \textbf{not} a BRA formula.  Guard's underline $\underline{(\cdot)}$
takes a value and returns its numeral term --- well-defined only at the
meta-level for closed numerical inputs.  For a variable $x$ (the case
Theorems~12--14 actually need), neither $\underline{x}$ nor
$\underline{f(x)}$ has any BRA referent: there is no operator in BRA's
term language that, given an arbitrary $\mathrm{Term}$, returns ``the
numeral of its value.''

The only formal object that \emph{does} correspond to Guard's
expression is its \emph{code}, the Term
\[
  \ulcorner f(\underline{x}) \;=\; \underline{f(x)} \urcorner,
\]
which Definition~12 of Guard's notes \emph{defines} by stipulation.  In
BRA2 this Term is built parametrically as $\mathit{codeFTeq}_1\;f\;x$,
with $\mathit{cor}\;x$ filling the slot for the code of $x$'s numeral
and $\mathit{cor}\;(f\,x)$ filling the slot for the code of $f(x)$'s
numeral.

When $x$ is a free Term variable, neither $\mathit{cor}\;x$ nor
$\mathit{cor}\;(f\,x)$ reduces (the defining equation
$\mathit{corOfReify}$ fires only on $\mathit{IsValue}$ inputs), so
$\mathit{codeFTeq}_1\;f\;x$ contains $\mathit{ap}_1$-headed subterms
literally.  Codes (in the sense of $\mathit{codeFormula}\;P$) are
$\mathit{IsValue}$-shape, so $\mathit{codeFTeq}_1\;f\;x$ at variable $x$
is \textbf{not} of the form $\mathit{codeFormula}\;P$ for any
$P : \mathit{Formula}$.

\subsection{Theorem 12 --- the encoded ``$\mathit{th}\;f$'' clause}

For every $\mathit{Fun}_1$ functor $f$, Theorem~12 produces a
$\mathit{Fun}_1$ $D_f$ such that $\mathit{thmT}$ evaluated at $D_f\;x$ is
provably equal to the encoded form of
``$f(\underline{x}) = \underline{f(x)}$'':

\begin{theorem}[$\mathit{thm12}\_F_1$]
\(
  \mathit{thm12}\_F_1 : (f : \mathit{Fun}_1) \to \mathit{Thm12\_F_1\_Result}\;f
\)
where $\mathit{Thm12\_F_1\_Result}\;f$ packages $D_f : \mathit{Fun}_1$ with
\(
  \mathit{proof} : (x : \mathit{Term}) \to
   \mathit{Deriv}\;\bigl(\mathit{thmT}\;(D_f\;x)
       \;=\; \mathit{codeFTeq}_1\;f\;x\bigr).
\)
\end{theorem}
The conclusion's RHS is the literal Pair-shape encoded equation.  An
analogous statement for $\mathit{Fun}_2$.  Structural recursion on $f$
with one clause per constructor (15 total),
BRA2/Thm12.agda module $\mathit{FromBridges}$.

\subsection{Theorem 13 --- under hypothesis $f\;x = y$}

\begin{lemma}[$\mathit{thm13}\_F_1$]
For every $f : \mathit{Fun}_1$ and every Theorem~12 result $r_{12}$ for $f$,
\(
  \mathit{thm13}\_F_1\;f\;r_{12}\;x\;y : \mathit{Deriv}\;(f\;x = y) \to
    \mathit{Deriv}\;(\mathit{thmT}\;(D_f\;x) = \mathit{codeFTeq}_{1,\mathrm{Hyp}}\;f\;x\;y).
\)
\end{lemma}
The conclusion replaces $\mathit{cor}\;(f\,x)$ by $\mathit{cor}\;y$.
One-line proof: $\mathit{ruleTrans}$ on $r_{12}.\mathit{proof}\;x$ with
$\mathit{cong}_R\;\mathit{Pair}$ on $\mathit{cong}_1\;\mathit{cor}$ of
the hypothesis.

\subsection{Theorem 14 --- the consistency-to-bot construction}

The contract abstracted in BRA2/Thm14Abstract.agda asks for:
\begin{enumerate}
  \item a $\mathit{Fun}_1$-functor $\mathit{constr5}$ such that
        $\mathit{thmT}\;(\mathit{constr5}\;x)$ syntactically encodes
        $\bot$;
  \item a parametric internal-implication
\[
  \mathit{step5} : (x : \mathit{Term}) \to
   \mathit{Deriv}\bigl((\mathit{thmT}\;x = \mathit{codeFormula}\;G)
                  \;\mathit{imp}\;
                  (\mathit{thmT}\;(\mathit{constr5}\;x) = \mathit{codeFormula}\;\bot)\bigr).
\]
\end{enumerate}

The closure $\mathit{closureToG}$ instantiates $\mathit{step5}$ at $v_1$,
$\mathit{ruleInst}$s $\mathit{Con}$, modus-ponenses, transports through
$\mathit{subIIeqcG}$ and $G_{\mathrm{unfold}}$, arriving at
$\mathit{Deriv}\;G$.  Composing with $\mathit{thm11}$ gives $\mathit{godelII}$.

\subsubsection{Construction of $\mathit{constr5}$}

Two pre-existing tautologies, encoded:
\begin{itemize}
  \item $t := (v_0 = v_2)\;\mathit{imp}\;((v_1 = v_2)\;\mathit{imp}\;(v_0 = v_1))$ (transitivity);
  \item $t' := (v_0 = v_1)\;\mathit{imp}\;((\neg(v_0 = v_1))\;\mathit{imp}\;\bot)$ (ex falso).
\end{itemize}
Both BRA-derivable (BRA2/Thm14Numerals.agda); $\mathit{encode}$ yields
value-shaped Terms $\mathit{tterm}, \mathit{t'term}$ certified by
$t\_term\_isValue$ / $t'\_term\_isValue$ (these certificates are
load-bearing for the chain --- see Section~\ref{sec:internal}).

The chain, paralleling Guard's page~17:
\begin{align*}
  f\_\mathit{part}(x) \;&:=\; \mathit{ruleInst}\;v_0\;(\mathit{th}\;\hat x)\;\big(\mathit{ruleInst}\;v_1\;(\mathit{sub}\;i\;i)\;(\mathit{ruleInst}\;v_2\;\mathit{cor}\;G\;\mathit{tterm})\big),\\
  g\_\mathit{part}(x) \;&:=\; \mathit{mp}\big(\mathit{mp}(f\_\mathit{part}(x), D_{\mathit{thmT}}\;x), D_{\mathit{sub}}\;i\;i\big),\\
  K\_\mathit{part}(x) \;&:=\; \mathit{ruleInst}\;v_1\;(\mathit{cor}\;x)\;x,\\
  h\_\mathit{part}\_\mathit{skel}(x) \;&:=\; \mathit{ruleInst}\;v_0\;(\mathit{th}\;\hat x)\;\big(\mathit{ruleInst}\;v_1\;(\mathit{sub}\;i\;i)\;\mathit{t'term}\big),\\
  \mathit{constr5}_{\mathrm{final}}(x) \;&:=\; \mathit{mp}\big(\mathit{mp}(h\_\mathit{part}\_\mathit{skel}(x), g\_\mathit{part}(x)), K\_\mathit{part}(x)\big).
\end{align*}

\subsubsection{The five steps}

(BRA2/Th14*.agda.)
\begin{description}
  \item[Step 1 --- $\mathit{step1\_l}$] Theorem~13 applied to $\mathit{thmT}$.
  \item[Step 2 --- $\mathit{step2\_l}$] Theorem~13 applied to $\mathit{sub}$ at $(j, j, \mathit{cor}\;G)$ under closed hypothesis $\mathit{subIIeqcG}$.
  \item[Step 3 --- $\mathit{step3\_l}$] Three sb-applications on $t$, canonicalised, then two $\mathit{mp}$ dispatches.
  \item[Step 4 --- $K\_\mathit{part}\_l$] One sb-application at $v_1$ on $G_{\mathrm{norm}}$ via the $\mathit{PrAtX}\;x$ hypothesis to swap $\mathit{thmT}\;x \mapsto \mathit{cor}\;G$.
  \item[Step 5 --- $\mathit{step5\_l}$] Two outer $\mathit{mp}$ dispatches at $\mathit{constr5}_{\mathrm{inner,final}}$ and $\mathit{constr5}_{\mathrm{final}}$.
\end{description}

\section{What is going on at the encoded layer: a remarkable internal proof}
\label{sec:internal}

\subsection{The phenomenon}

The chain $\mathit{step1\_l},\dots,\mathit{step5\_l}$ proves
$\mathit{Deriv}$s of the schematic form
\[
  \mathit{Deriv}\;\bigl(\mathit{PrAtX}\;x \;\mathit{imp}\;
    \mathit{atomic}\;(\mathit{eqn}\;(\mathit{thmT}\;t_i)\;u_i)\bigr),
\]
where $t_i, u_i$ are Terms depending on $x$.  Several of these
intermediate $u_i$ are \textbf{not} of the form $\mathit{codeFormula}\;P$
for any $P : \mathit{Formula}$.

A representative example: $\mathit{step3\_l}$ (BRA2/Th14Step3.agda)
concludes
\[
  \mathit{Deriv}\;\bigl(\mathit{PrAtX}\;x \;\mathit{imp}\;
    \mathit{atomic}\;(\mathit{eqn}\;(\mathit{thmT}\;(g\_\mathit{part}\;x))\;
      \mathit{ap}_2\;\mathit{Pair}\;(\widehat{\mathit{th}\;\hat x})\;\widehat{\mathit{sub}\;i\;i})\bigr),
\]
with
\[
  \widehat{\mathit{th}\;\hat x}
   \;=\; \mathit{ap}_2\;\mathit{Pair}\;\mathit{tagAp}_1\;
        (\mathit{ap}_2\;\mathit{Pair}\;(\mathit{codeF}_1\;\mathit{thmT})\;(\mathit{ap}_1\;\mathit{cor}\;x)).
\]
At a variable $x$ (e.g.\ $x = v_1$), $\mathit{ap}_1\;\mathit{cor}\;x$
does not reduce: $\mathit{corOfReify}$ requires $\mathit{IsValue}\;x$,
and $v_1$ is not value-shape.  The Term
$\mathit{ap}_1\;\mathit{cor}\;x$ has an $\mathit{ap}_1$ head; codes
are exclusively $\mathit{IsValue}$-shape (built from $O$ and
$\mathit{ap}_2\;\mathit{Pair}$).  Therefore $u_3 = \mathit{ap}_2\;
\mathit{Pair}\;(\widehat{\mathit{th}\;\hat x})\;\widehat{\mathit{sub}\;i\;i}$
is not in the image of $\mathit{codeFormula}$.

The same phenomenon recurs in $\mathit{step4}, \mathit{step5}_{\mathrm{intermediate}},
h\_\mathit{part}\_\mathit{skel}\_l$.  In each case, $u_i$ is a
\emph{substituted-code} --- the result of inserting $\mathit{cor}\;x$
into a variable slot of the code of an open formula.

Yet $\mathit{Deriv}\;(\mathit{atomic}\;(\mathit{eqn}\;t\;u))$ is
perfectly well-formed for arbitrary Terms $t, u$ --- the equational
predicate $\mathit{eqn}$ does not require $u$ to be a code of a formula.
The chain composes via the standard inference rules ($\mathit{mp}$,
$\mathit{ruleTrans}$, $\mathit{cong}_*$, $\mathit{axS}$, $\mathit{axK}$)
without ever needing the intermediate $u_i$ to be $\mathit{codeFormula}\;P$.

\subsection{Why this works: substituted-codes vs.\ codes-of-formulas}

There are two distinct readings of Guard's notation
$\ulcorner f(\underline{x}) \;=\; \underline{f(x)} \urcorner$:

\begin{description}
  \item[(A) Code-of-substituted-formula.]  Read $\underline{x}$ as
    $\mathit{cor}\;x$ (a BRA Term), and form the BRA formula
    $\mathit{atomic}\;(\mathit{eqn}\;(\mathit{ap}_1\;f\;(\mathit{cor}\;x))\;\dots)$.
    Take its $\mathit{codeFormula}$.
    The inner slot becomes $\mathit{code}\;(\mathit{cor}\;x)$
    (= the code of the BRA term $\mathit{cor}\;x$).
  \item[(B) Substituted-code.]  Take the open formula
    $\mathit{atomic}\;(\mathit{eqn}\;(\mathit{ap}_1\;f\;v_n)\;\dots)$,
    code it, then $\mathit{sb}$-substitute $\mathit{cor}\;x$ into the
    $\ulcorner v_n\urcorner$ slot.  The inner slot becomes
    $\mathit{cor}\;x$ itself.
\end{description}
At a free variable $x$ these readings disagree: (A) places
$\mathit{code}\;(\mathit{cor}\;x)$ in the slot, (B) places
$\mathit{cor}\;x$.  Theorem~12's $\mathit{codeFTeq}_1$ uses reading
(B) --- the substituted-code form.  This is forced by the structure of
$D_f$, which is built from $\mathit{Comp}_2\;\mathit{Pair}\;\dots\;\mathit{cor}$,
producing $\mathit{cor}\;x$ in its evaluation directly.

The two readings coincide \textbf{when $x$ is value-shaped}:
$\mathit{corOfReify}$ then gives
$\mathit{cor}\;x = \mathit{code}\;x$, and \emph{Lemma~codeCommutesFormula}
(BRA2/CodeCommutes.agda) gives
\[
  \mathit{code}\;(\mathit{substF}\;n\;t\;P) \;=\;
   \mathit{codedSubst}\;(\mathit{code}\;t)\;(\mathit{code}\;(\mathit{var}\;n))\;(\mathit{codeFormula}\;P).
\]
So at value-shape $x$, the substituted-code (B) is BRA-equal to
$\mathit{codeFormula}\;(\mathit{substF}\;n\;x\;P_{\mathrm{open}})$ ---
a genuine code-of-formula.  At variable $x$ this collapse fails, and
(B) remains a hybrid Term.

\subsection{The chain's discharge mechanism}

Why does Goedel~II close even though step~3 etc.\ produce hybrid RHS?

\begin{enumerate}
  \item \emph{Internal composition stays at the equational level.}
        The chain's mp/cong/ruleTrans steps operate on $\mathit{eqn}\;t\;u$
        for arbitrary $t, u$; they do not require $u$ to be
        $\mathit{codeFormula}\;P$.  All intermediate Derivs are honest
        BRA derivations producing well-typed atomic equations.
  \item \emph{$\mathit{step5}$'s conclusion is genuinely
        $\mathit{codeFormula}\;\bot$.}  At step~5 the final RHS is
        $\mathit{codeFormula}\;\bot$, where $\bot$ is closed.  No
        $\mathit{cor}\;x$ remains in the conclusion --- the
        Comp2-of-KT-of-codeFormula-bot construction in
        $\mathit{constr5}_{\mathrm{final}}$ is a constant in $x$
        modulo Lift, so its $\mathit{thmT}$-image is a closed code by
        construction.
  \item \emph{Closure transports through $\mathit{notEqTransport}$ at
        a value-shape witness.}  $\mathit{closureToG}$ instantiates
        $\mathit{step5}$ via $\mathit{ruleInst}$ at the encode-witness
        of $G$'s proof.  By the strengthened $\mathit{encode}$ contract
        (returning $\mathit{IsValue}$), this witness is a value-shaped
        Term.  At such a witness, $\mathit{cor}\;x$ inside any
        intermediate substituted-code rewrites to $\mathit{code}\;x$ via
        $\mathit{corOfReify}$, and the substituted-codes collapse to
        genuine $\mathit{codeFormula}$s of substituted formulas.
\end{enumerate}

The interleaving --- carry hybrid forms through internal-implication
chains, collapse at the closure --- is what lets the proof be
parametric over $(x : \mathit{Term})$ rather than meta-iterated over
closed numerals.

\subsection{The projective specification of $\mathit{mp}$}
\label{subsec:mp-projective-spec}

A natural question raised by the chain is: what \emph{exactly} is the
specification of $\mathit{mp}$ as embedded in $\mathit{thmT}$, given
that step~3 applies $\mathit{mp}$-internalisation to inputs that are
not $\mathit{codeFormula}\,(P \mathrel{\mathit{imp}} Q)$ for any
$P, Q : \mathit{Formula}$?

A naive declarative reading would assert
\[
  \mathit{mp}\big(\mathit{codeFormula}\,(P \mathrel{\mathit{imp}} Q),
                  \mathit{codeFormula}\,P\big)
   \;=\; \mathit{codeFormula}\,Q.
\]
This is the soundness reading at well-typed inputs.  But it is
\emph{not} what BRA's $\mathit{thmT}$ actually computes, and it is
not what step~3 uses.  The actual specification is the
verifying-body cascade in BRA2/SoundMpVProof.agda:

\begin{quote}
  Given a coded proof tree
  $\mathit{arg} = \mathit{ap}_2\;\mathit{Pair}\;\mathit{tagMp}\;
   (\mathit{ap}_2\;\mathit{Pair}\;y_1\;y_2)$,
  $\mathit{thmT}(\mathit{arg})$ dispatches via
  $\mathit{body\_mp\_v}(a, b) = \mathit{Post}\;\mathit{verifierF}_1\;\mathit{Pair}\;a\;b$
  where $\mathit{verifierF}_1$ first computes
  $\mathit{thmT}\,y_1$ and $\mathit{thmT}\,y_2$, then performs
  \emph{two} $\mathit{TreeEq}$ checks:
  \begin{enumerate}
    \item $\mathit{outerCheck}$:
          $\mathit{Fst}(\mathit{thmT}\,y_1) \stackrel{?}{=} \mathit{tagImp}$,
    \item $\mathit{innerCheck}$:
          $\mathit{thmT}\,y_2 \stackrel{?}{=} \mathit{Fst}(\mathit{Snd}(\mathit{thmT}\,y_1))$.
  \end{enumerate}
  If both succeed, the verifier returns $\mathit{Snd}(\mathit{Snd}
  (\mathit{thmT}\,y_1))$, the consequent slot.  If either fails, it
  returns $\mathit{codeTriv} = \mathit{falseT}$, the safe default.
\end{quote}

\paragraph{Soundness.}  The two $\mathit{TreeEq}$ checks together
ensure that the consequent extracted is the consequent of \emph{this
specific implication} applied to \emph{this specific antecedent}.
Without the inner check $\mathit{thmT}\,y_2 = \mathit{Fst}(\mathit{Snd}
(\mathit{thmT}\,y_1))$, BRA's $\mathit{mp}$-dispatch would project
the second slot of an arbitrary $\mathit{tagImp}$-headed Term, with
no relation to $y_2$, and $\mathit{thmT}$ would fail to be a sound
proof-checker (the encoding-soundness lemma in
BRA2/Thm/ThmT.agda's $\mathit{encodeRich}$ would not go through).
The inner check is what makes the $\mathit{mp}$ clause of
$\mathit{encodeRich}$ valid: it certifies, equationally inside BRA,
that the two sub-proofs $y_1, y_2$ are compatible at the
expected antecedent.

\paragraph{Why this still works on substituted-codes.}  The two
$\mathit{TreeEq}$ checks are \emph{equational comparisons on Terms},
not type-checks on Formulas.  Whatever sits in the inner slots of
$\mathit{thmT}\,y_1$ and $\mathit{thmT}\,y_2$ --- $\mathit{codeFormula}\,P$
and $\mathit{codeFormula}\,Q$, or $\mathit{cor}\,x$ interleaved with
tag constants, or any value-shaped Term --- only matters in that
the inner check fires correctly: the antecedent slot of
$\mathit{thmT}\,y_1$ must be $\mathit{TreeEq}$-equal to
$\mathit{thmT}\,y_2$.

In step~3 of Theorem~14, the equational chain is set up precisely so
that this match holds: each $\mathit{mp}$-internalised step ensures,
via the previous step's $\mathit{Deriv}$, that the antecedent slot of
the current implication's substituted-code is exactly the consequent
of the previous step.  So the two $\mathit{TreeEq}$ checks both fire
\emph{equationally}, with both sides being substituted-codes, and
the dispatch returns the consequent's substituted-code.  The chain
composes precisely because both readings of $\mathit{mp}$ --- the
``well-typed Formula'' and ``well-shaped substituted-code'' cases ---
are handled by the same projective primitive-recursive computation.

The result is again a substituted-code, fed into the next step the
same way:
\[
  \mathit{mp\text{-}body}\big(\text{substituted-code-of-implication},
                              \text{substituted-code-of-antecedent}\big)
   \;=\; \text{substituted-code-of-consequent},
\]
equationally, with no claim about Formula content but with the
soundness-relevant antecedent-match certified by $\mathit{TreeEq}$.

The two readings of $\mathit{mp}$'s specification stand in a clear
hierarchy:
\begin{description}
  \item[Projective (what BRA actually computes).]  The $\mathit{Fst}$/
        $\mathit{Snd}$ projection above, parametric in the inner
        Terms.  Total on all Term inputs of the right outer shape,
        producing a Term output by syntactic projection.
  \item[Declarative (what soundness justifies).]  At literal
        $\mathit{codeFormula}\,(P \mathrel{\mathit{imp}} Q)$ inputs, the
        projective behaviour delivers $\mathit{codeFormula}\,Q$.  This
        is what makes $\mathit{thmT}$ a sound proof-checker
        ($\mathit{encodeRich}$'s $\mathit{mp}$ clause).
\end{description}
The soundness reading factors through the projective reading: BRA's
$\mathit{mp}$ primitive only knows about Term shape, and Formula-content
correctness is the \emph{consequence} for the well-shaped subset of
inputs.  Theorem~14 step~3 uses the projective reading directly; the
final closure converts to the declarative reading by instantiating at
a value-shape witness.

The same observation lifts to the other ``soundified'' rule
dispatchers in the chain ($\mathit{ruleSym}$, $\mathit{cong}_1$,
$\mathit{cong}_L$, $\mathit{cong}_R$, $\mathit{ruleTrans}$,
$\mathit{ruleInst}$, $\mathit{ruleInst}_2$, $\mathit{ruleIndBT}$).
Each is specified as a primitive-recursive Term projection on a
shape-discriminated input; each delivers a Formula-content
interpretation only on its well-shaped subset.  The
verifying-body construction
($\mathit{body\_X\_v} = \mathit{Post}\,\mathit{verifierX}\,\mathit{Pair}$,
Section~``Sound $\mathit{thmT}$'' of \verb|README.md|) is precisely
what enforces the safe default $\mathit{codeTriv}$ on malformed inputs
without spoiling the projective behaviour on well-shaped ones.

In one line: \emph{BRA's $\mathit{mp}$ is a Term projection that
happens to internalise Formula modus ponens at well-typed inputs}.
The chain in step~3 uses the projection-on-Term reading; the
internal-modus-ponens reading is what justifies it semantically at
the value-shape closure.  Both readings are valid specifications;
the projective one is strictly more general and is what makes the
chain compose at variable $x$.

\subsection{Comparison with Hilbert--Bernays--L\"ob}

Standard textbook G\"odel~II proofs go through provability statements:
\[
  \text{D1} : \vdash A \;\Longrightarrow\; \vdash \mathrm{Pr}(\ulcorner A\urcorner),
  \quad
  \text{D2} : \vdash \mathrm{Pr}(\ulcorner A\to B\urcorner)
              \to \mathrm{Pr}(\ulcorner A\urcorner)
              \to \mathrm{Pr}(\ulcorner B\urcorner),
  \quad
  \text{D3} : \vdash \mathrm{Pr}(\ulcorner A\urcorner) \to
              \mathrm{Pr}(\ulcorner \mathrm{Pr}(\ulcorner A\urcorner)\urcorner).
\]
Every premise is a provability statement \emph{about} a code of a
specific formula.  The chain stays at the level ``$F$ is provable'' for
various $F$.  Reflection (D3) is needed to handle the iteration.

Guard's BRA chain operates one level lower: at the level of
\emph{equations} $\mathit{thmT}\;t = u$ for arbitrary Terms $t, u$.
Reflection (D3-analogue) is realised internally and parametrically,
without reference to any specific formula:
\begin{itemize}
  \item Theorem~12 is the parametric ``$D_f$ is the
        encoded image of $f$'' --- one statement, all $f$.
  \item Theorem~13 lifts external equations to encoded form,
        parametric in the equational hypothesis.
  \item Theorem~14 chains substituted-codes through $\mathit{mp}$/
        $\mathit{sb}$/$\mathit{ruleInst}$, never invoking a separate
        ``$\mathrm{Pr}(\ulcorner \mathrm{Pr}(\ulcorner G\urcorner)\urcorner)$''
        reflection axiom.  $\mathit{step5}$ is an equational
        consequence of Theorem~12's defining equations under hypothesis,
        not a provability claim about a specific formula.
\end{itemize}

\subsection{Remarkable consequences}

\begin{enumerate}
  \item \textbf{Avoids meta-reflection.}  No $\mathrm{Pr}(\ulcorner
        \mathrm{Pr}(\ulcorner F\urcorner)\urcorner)$ axiom is invoked.
        The analogous internalisation falls out of $\mathit{thmT}$'s
        defining equations + Comp2/KT/cor compositions.
  \item \textbf{Fully equational.}  Goedel~II reduces to BRA equational
        reasoning + $\mathit{mp}$/$\mathit{axS}$/$\mathit{axK}$/
        $\mathit{axNeg}$/$\mathit{axContrapos}$.  No specialised
        ``provability logic'' is needed.
  \item \textbf{Parametric over $(x : \mathit{Term})$.}  Each step lemma
        works for any Term $x$ (variable, application of cor to a
        variable, anything).  Standard HBL proofs are usually
        meta-iterated over each closed witness.
  \item \textbf{The intermediate Derivs are genuine BRA proofs that do
        not correspond to ``BRA proves formula $F$'' for any single
        $F$.}  This is the structurally interesting feature: the chain
        proves \emph{equational} statements between Terms, with the
        ``provability'' interpretation recovered only at the closure.
\end{enumerate}

The cost is bookkeeping that the development makes explicit: the
substituted-code RHS at variable $x$ requires $\mathit{IsValue}$
certificates (on $j$, on encode-witnesses, on $t\_term$/$t'\_term$)
propagated through to the closure.  These certificates are the formal
counterpart of Guard's silent assumption that ``$\underline{x}$'' for variable
$x$ is sensible because $x$ will eventually be instantiated at a
closed numeral.

In short: \emph{Goedel~II in BRA is an equational proof at the code
layer, internalised via primitive recursion, with the
provability-logic story emerging only at closure}.  Guard's 1963
formulation has this character intrinsically; HBL-style proofs do not.
The Agda dev makes the distinction visible.

\section{Methodology}

\subsection{Constraints maintained throughout}

\begin{itemize}
  \item \verb|--safe --without-K --exact-split| on every file.
  \item ASCII only.  No Unicode.
  \item No postulates.  No holes.  No \verb|with|-abstraction.  No dot patterns
        (one exception: \verb|.O| / \verb|.a .b| in $\mathit{IsValue}$ pattern matches,
        which are the canonical syntax for matching against an indexed family's
        index).
  \item No new $\mathit{Deriv}$ axioms beyond Guard's primitive list.
\end{itemize}

\subsection{Verification}

\begin{verbatim}
$ agda --safe BRA2/GoedelIIFull.agda
$ echo $?
0
$ ls BRA2/GoedelIIFull.agdai
BRA2/GoedelIIFull.agdai
$ grep -rn "^postulate" BRA2/
$ # (empty)
\end{verbatim}

\section{Discussion}

The two G\"odel theorems in BRA2:
\begin{itemize}
  \item \textbf{G\"odel I.} If the diagonal sentence $G$ is provable, then $\bot$ is provable.
  \item \textbf{G\"odel II.} If the consistency formula $\mathit{Con}$ is provable, then $\bot$ is provable.
\end{itemize}

The intermediate functions $\mathit{constr5}_{\mathrm{final}}$ and the
parametric step lemmas $\mathit{step1\_l}$--$\mathit{step5\_l}$,
$K\_\mathit{part}\_l$, $h\_\mathit{part}\_\mathit{skel}\_l$ are
\emph{first-order} in the sense of recursive-arithmetic definability:
each is a closed combinator built from primitive $\mathit{Fun}_1$ /
$\mathit{Fun}_2$ symbols, and each $\mathit{Deriv}$ derivation is given
by an explicit equational proof in the BRA system.  No external
soundness theorem is invoked.

\subsection{Complexity calibration: ramified-elementary fragment}
\label{subsec:elementary}

Although BRA's term language admits unrestricted primitive recursion on
binary trees (the $\mathit{treeRec}$ constructor of $\mathit{Fun}_2$),
the entire chain reachable from
\verb|godelII : Deriv Con -> Deriv bot| in BRA2/GoedelIIFull.agda lies
within the \emph{ramified-elementary} fragment in the sense of
Leivant~\cite{leivant1999ramified}.  Specifically, every functor used
in the closure is well-typed under a stratification
$\Omega^k\theta \to \theta$ with $k \le 2$, where $\theta$ is the base
type of binary-tree-shaped Terms and $\Omega\theta$ is the type of
data supporting recurrence over $\theta$ in Leivant's discipline
(paper~III, Section~2.3, equation~(4)).

\paragraph{Inventory of recurrence-using functors.}
A grep over the closure confirms that across
\verb|Cor.agda|, \verb|SubT.agda|, \verb|Sb.agda|, \verb|Thm/ThmT.agda|,
\verb|Thm14Constr5Final.agda|, \verb|Th14Step5.agda|, and
\verb|Th12TreeRecInternal.agda|, the constructor $\mathit{treeRec}$
appears in exactly one syntactic definition
(\verb|Th12TreeRecInternal.agda:396|), plus the derived combinators
$\mathit{Rec}\;s = \mathit{Comp}_2\;(\mathit{treeRec}\;Z\;s)\;Z\;I$ and
$\mathit{RecP}\;s = \mathit{treeRec}\;Z\;s$.  The headline functors
unfold as follows.
\begin{itemize}
  \item $\mathit{cor} = \mathit{Rec}\;\mathit{stepCor}$: one
        $\mathit{treeRec}$, level $\Omega\theta \to \theta$.
        $\mathit{stepCor}$ is a $\mathit{Fan}/\mathit{Lift}/\mathit{KT}/\mathit{Post}/\mathit{Pair}$
        composite, no recurrence.
  \item $\mathit{subT} = \mathit{RecP}\;\mathit{stepSubT}$: one
        $\mathit{treeRec}$, level
        $\theta \times \Omega\theta \to \theta$.  $\mathit{stepSubT}$
        uses only $\mathit{Fan}$, $\mathit{IfLf}$, $\mathit{Lift}$,
        $\mathit{Post}$, $\mathit{Comp}$.
  \item $\mathit{sb} = \mathit{Fan}\;\ldots\;\mathit{subT}$: a single
        $\mathit{Fan}$ wrapping $\mathit{subT}$, level
        $\theta \times \Omega\theta \to \theta$ via $\mathit{subT}$.
  \item $\mathit{thmT} = \mathit{Rec}\;\mathit{stepProtoWrapped}$:
        one $\mathit{treeRec}$, level $\Omega\theta \to \theta$.
        $\mathit{stepProtoWrapped}$ is the 40-tag
        $\mathit{IfLf}$/$\mathit{Fan}$/$\mathit{Lift}$/$\mathit{KT}$/$\mathit{TreeEq}$
        cascade, no recurrence.
  \item $D_{\mathit{thmT}} = \mathit{thm12\_F1}\;\mathit{thmT}.\mathit{Df}$:
        depth-2 use of $\mathit{treeRec}$.  The outer
        $\mathit{treeRec}\;\mathit{lf\_emitter}\;\mathit{step\_emitter}$
        walks the input encoding; inside $\mathit{step\_emitter}$, the
        original $\mathit{treeRec}\;f\;s$ is invoked as
        $\mathit{Comp}_2\;(\mathit{treeRec}\;f\;s)\;\mathit{Fst}_1\;v_1^{\mathit{Fun1}}$,
        i.e.\ on a $\mathit{Fst}$/$\mathit{Snd}$-extracted
        \emph{critical sub-argument} of the outer recurrence argument
        (BRA2/Th12TreeRecInternal.agda:222--226, 354--356).  This is
        Leivant's allowed pattern: the critical args $\vec a$ in
        $g_{c_i}(\vec x)(\vec a)(\ldots)$ live at type $\Omega\theta$
        when the outer recurrence argument is at $\Omega\theta$, so
        re-invoking $f : \ldots, \Omega\theta \to \theta$ on them is
        well-typed.  Level: $\theta \times \Omega^2\theta \to \theta$
        when seen one level higher; safely $\Omega\theta \to \theta$
        when applied with the outer recurrence argument at
        $\Omega^2\theta$.
  \item $D_{\mathit{sub\_at\_ii}}$, $g_{\mathit{part}}$,
        $h_{\mathit{part}}^{\mathit{skel}}$, $K_{\mathit{part}}$,
        $\mathit{constr}_5^{\mathit{final}}$: $\mathit{Comp}_2$-of-$\mathit{Pair}$
        / $\mathit{KT}$ compositions of the above.  No new
        $\mathit{treeRec}$.
\end{itemize}

\paragraph{Maximum nesting depth: 2.}
Across the entire closure, no $\mathit{treeRec}$ ever takes its
recurrence argument from a recursion \emph{result} of another
$\mathit{treeRec}$.  The single nested case
($D_{\mathit{thmT}}$) is the structural-recursion-on-critical-arg form,
which is well-typed at level $\Omega^2\theta \to \theta$ in Leivant's
discipline.  Every functor in the closure terminates in time
elementary in the size of its input.

\paragraph{Implication: G\"odel~II in elementary BRA.}
By Leivant's main theorem (Paper~III, Theorem at p.\,210), the
functions definable by ramified higher-type recurrence at types of
order at most $k$ are exactly the elementary (Kalmar) functions
$\mathcal{E}_3$.  The BRA2 closure thus computes only elementary
functions on Tree-codes.  This is sharp: by Wilkie--Paris and Buss's
formalisation~\cite{wilkieparis1987}, elementary arithmetic
$\mathit{EFA} = I\Delta_0 + \mathrm{EXP}$ already proves the second
incompleteness theorem; nothing weaker than $\mathrm{EXP}$ suffices,
because Hilbert--Bernays--L\"ob condition~D3 ($\Sigma_1$-completeness
on closed numerical statements) requires at least exponential
encoding-arithmetic.  BRA2 sits at exactly this threshold: enough to
encode and substitute proofs, no stronger.  The headline result of
the development can therefore be sharpened from ``G\"odel~II in BRA''
to:
\begin{quote}
\textbf{Theorem (calibrated headline).}  G\"odel's second
incompleteness theorem is provable in the ramified-elementary
fragment of BRA --- the $\Omega^k\theta \to \theta$ fragment with
$k \le 2$ in Leivant's higher-type ramification --- with no use of
unrestricted higher-type primitive recursion.
\end{quote}
This calibration is a post-hoc certificate: it does not require any
change to the existing \verb|--safe| Agda development, only the
observation that none of the constructed functors exit the elementary
fragment.  The audit (file inventory, $\mathit{treeRec}$ occurrence
count, depth-2 nesting check) is a finite verification on the seven
files listed above.

\subsection{Induction calibration: atomic-predicate tree induction only}
\label{subsec:atomic-ind}

The function-class audit of Section~\ref{subsec:elementary} measures
\emph{which functors} are needed.  A complementary audit measures
\emph{which induction rules and at what predicate complexity} are
needed.  This second audit is even sharper.

BRA's primitive induction rule
$\mathit{ruleIndBT}$ (and its diagonal variant
$\mathit{ruleIndBT}_2$, used for $\mathit{TreeEq}$'s pair-of-pairs
recursion; see BRA2/Deriv.agda lines~218--253) accepts an
\emph{arbitrary} $\mathit{Formula}$ as the induction predicate ---
atomic equations, negations, implications, and any nesting thereof.

\paragraph{Inventory of induction call sites.}
A clean recheck of \verb|BRA2/GoedelIIFull.agda| under
\verb|--safe| (147 files in the closure) lists exactly three
$\mathit{ruleIndBT}$ / $\mathit{ruleIndBT}_2$ call sites, with the
following induction predicates.
\begin{enumerate}
  \item \verb|TreeEqReflParam.agda:158|.
        $\mathit{motiveTreeEqRefl}
        = \mathit{atomic}\;(\mathit{eqn}\;(\mathit{ap}_2\;\mathit{TreeEq}\;(\mathit{var}\;0)\;(\mathit{var}\;0))\;O)$.
        \emph{Atomic equation.}
  \item \verb|Th12TreeEqUniv.agda:72| ($\mathit{ruleIndBT}_2$).
        $P_{\mathit{Th12\_TreeEq}}
        = \mathit{atomic}\;(\mathit{eqn}\;
              (\mathit{ap}_1\;\mathit{thmT}\;(\mathit{ap}_2\;D_{\mathit{TreeEq}}\;
                  (\mathit{var}\;0)\;(\mathit{var}\;1)))\;
              \mathit{codeFTeq}_2^{\mathit{TreeEq}}\;\ldots)$.
        \emph{Atomic equation.}
  \item \verb|Th12TreeRecInternal.agda:1786|.
        $P_{\mathit{Th12\_treeRec}}
        = \mathit{atomic}\;(\mathit{eqn}\;
              (\mathit{ap}_1\;\mathit{thmT}\;(\mathit{ap}_2\;D_2^{\mathit{treeRec}_{f,s}}\;
                  (\mathit{var}\;1)\;(\mathit{var}\;0)))\;
              \mathit{codeFTeq}_2^{\mathit{treeRec}_{f,s}}\;\ldots)$.
        \emph{Atomic equation.}
\end{enumerate}

\textbf{Every induction predicate in the closure is an atomic
equation.}  No induction over an implication, no induction over a
negation, no induction over a compound boolean combination, no
nested predicate of any kind.  Restricting BRA's
$\mathit{ruleIndBT}$ to the atomic-predicate fragment ---
$\mathit{IndBT\text{-}atomic}$ --- the entire BRA2 proof of
Goedel~II survives unchanged.

\paragraph{Significance.}
This is the BRA-language analog of \emph{open induction}
(Shepherdson's $\mathrm{IOpen}$) / \emph{quantifier-free induction},
specialised to BRA's vocabulary of primitive recursive functors.
Since BRA has no quantifiers in its formula language at all, and the
induction predicates here are not even disjunctions or implications
but single equations between primitive-recursive Term-expressions,
the relevant fragment sits \emph{strictly below} EFA = $I\Delta_0 +
\mathrm{EXP}$ (which has $\Delta_0$-induction with bounded
quantifiers) and considerably below PRA (full quantifier-free
induction over arbitrary boolean combinations).

\paragraph{Caveat: step rule still uses propositional logic.}
While the induction \emph{predicate} is atomic in all three sites,
the induction \emph{step} hypothesis fed to $\mathit{ruleIndBT}$
has shape
\[
  P[O/v_1] \;\supset\; (P[O/v_2] \;\supset\; P[\mathit{Pair}(v_1, v_2)/0]),
\]
an implication of three substitution-instances of the atomic
predicate.  Building and discharging it requires
$\mathit{mp}$ and the propositional axioms ($\mathit{axK}$,
$\mathit{axS}$).  Furthermore, the negation closure around
$\mathit{Con}$ (which has shape
$\mathit{not}\;(\mathit{atomic}\;\ldots)$) requires classical
propositional axioms: $\mathit{axContrapos}$, $\mathit{axExFalso}$,
$\mathit{axNeg}$, used in
\verb|Thm11Diagonal.agda|, \verb|Thm14Abstract.agda|,
\verb|Thm11.agda|, and \verb|Thm14Numerals.agda|.  These are not
induction principles; they are classical propositional logic.

\paragraph{Combined calibrated headline.}
The two complementary audits give the following sharpened statement
of the BRA2 result:
\begin{quote}
\textbf{Theorem (calibrated).}  G\"odel's second incompleteness
theorem is provable in BRA using only:
\begin{itemize}
  \item \emph{Functions:} the ramified-elementary fragment, $\Omega^k\theta \to \theta$
        with $k \le 2$ in Leivant's higher-type ramification --- equivalent
        to Kalmar's elementary class $\mathcal{E}_3$;
  \item \emph{Induction:} tree induction at atomic-equation predicates
        only --- the BRA analog of open / quantifier-free induction;
  \item \emph{Logic:} classical propositional logic (modus ponens,
        $\mathit{axK}$, $\mathit{axS}$, $\mathit{axContrapos}$,
        $\mathit{axExFalso}$, $\mathit{axNeg}$), equational
        congruence, and Hilbert-style instantiation.
\end{itemize}
No use of higher-type primitive recursion beyond level~2; no
induction over compound formulas; no quantifiers.
\end{quote}
The two fragment certificates are independent: the first restricts
the function vocabulary (depth of $\mathit{treeRec}$ nesting), the
second restricts the predicate vocabulary (formula complexity at the
induction site).  Both are post-hoc certificates, observable by a
finite \verb|grep| audit of the seven function-defining files plus
the three induction call sites; no Agda re-proof is required.

\subsection{Comparison with prior formalisations}

The only previous machine formalisation of G\"odel's \emph{Second}
Incompleteness Theorem we are aware of is Paulson's Isabelle/HOL
development of hereditarily finite set theory~\cite{paulson2014},
following \'{S}wierczkowski~\cite{swierczkowski2003}.  Other
formalisations (O'Connor in Coq, Han in Lean, Carneiro's metamath
work) target G\"odel's \emph{First} theorem only.

Paulson's development uses HF set theory and full G\"odel numbering of
Hilbert-style proofs, with encoded substitution treated as an external
soundness theorem.  Bound variables in the formal language (HF has
$\exists$ and $\forall$ as primitive binders, with bound variables
significant for inductive proofs over formula structure) are handled
via \emph{Nominal Isabelle}~\cite{urbankaliszyk2012}: a non-standard
extension of Isabelle/HOL that quotients binding constructions
modulo $\alpha$-equivalence, with permutations on names as the
underlying mechanism.  Paulson reports that the nominal framework
is essential to make induction over formula structure tractable
(O'Connor's first-incompleteness Coq formalisation needed roughly
1900 lines for a single substitution lemma without nominal support);
the cost is that any function defined on formulas must be proved to
use bound variables ``sensibly,'' a property whose proofs are
demanding and required dedicated assistance from Christian Urban.
For coding, Paulson separately uses de~Bruijn indices and proves a
correspondence with the nominal syntax.

The present development sidesteps this entire issue.  BRA's
$\mathit{Formula}$ type has no quantifiers --- only atomic equations,
$\neg$, and $\supset$ --- so there are no bound variables in the
formal language at all; what appears in $\mathit{atomic}\;
(\mathit{eqn}\;\ell\;r)$ are just object-level Terms, with free
variables denoted $\mathit{var}\;n$ for $n : \mathbb N$.  Universal
closure of an open formula is expressed implicitly via the BRA rule
$\mathit{ruleInst}$ (substitution into a derived formula), without
any meta-level renaming.  Substitution at the term level is the
ordinary primitive-recursive $\mathit{subst} : \mathbb N \to
\mathit{Term} \to \mathit{Term} \to \mathit{Term}$, capture-free by
construction (no binder under which to capture).  Coded substitution
$\mathit{sub} : \mathit{Fun}_2$ is itself a BRA functor; its
correctness is the equational $\mathit{codeCommutesFormula}$ lemma,
not a metatheoretic substitution-lemma proof.  In short, what
required Nominal Isabelle in Paulson's HF formalisation is for us a
non-issue, because Guard's BRA was designed without binding
constructors in the first place.

\paragraph{Paulson's ``mistake'' in Boolos--\'{S}wierczkowski, and
Guard's analogous construction.}
Paulson devotes \S6 of his paper to identifying a presentational
flaw in Boolos~\cite{boolos1993} and \'{S}wierczkowski's expositions
of the second incompleteness theorem in HF.  Both sources describe
the internal substitution as $\mathrm{Pf}\lfloor\alpha\rfloor_V(Q)$,
read informally as ``$\alpha$ with each free variable $x_i$ replaced
by the formal term $Q(x_i)$,'' where $Q$ is a ``pseudo-function''
mapping a value to its name in the formal language.  Paulson's
critique: $Q$ is not a function of the formal language (HF has only
one function symbol, $\mathord{\lhd}$), and any version of the
substitution that leaves the original variables $x$ free in the
transformed formula does not work --- because $x$ ranges over all
HF sets, including codes of formulas, and the supposed equation
$Q(x) = $ ``the numeral of $x$'' has no formal status.  Paulson's
corrected version introduces \emph{new} free variables $x'$ that
appear in the transformed formula, constrained by relational
antecedents $\mathrm{QR}(x, x')$ on the left of the turnstile.  In
\S7 he further shows that \'{S}wierczkowski's earlier attempt to
define such a $Q$ via a total ordering on the HF universe has a
significant gap, which Paulson closes by working with the relation
$\mathrm{QR}$ throughout instead of the function $Q$.

The same conceptual issue arises in Guard's notation in
\verb|guard15.pdf|.  Guard writes the encoded equational defining
clause for a primitive functor $f$ as
\[
  \mathit{th}(D_f\,x) \;=\; \ulcorner f(\underline{x}) \;=\; \underline{f(x)}\urcorner,
\]
where the underline $\underline{(\cdot)}$ takes a value to its
numeral term --- exactly Boolos--\'{S}wierczkowski's pseudo-function
$Q$ at a different name.  At a free variable $x$ this notation is
not a BRA expression: there is no operator in BRA's term language
that, given an arbitrary $\mathrm{Term}$, returns ``the numeral of
its value'' (Section~\ref{subsec:underline-caveat}).  Read
literally, Guard's $\underline{x}$ would be the same kind of
pseudo-function Paulson criticises.

The BRA chain treats this rigorously the way Paulson treats his
$\mathrm{QR}$.  The combinator $\mathit{cor} : \mathit{Fun}_1$ is a
genuine BRA functor (Layer~3, BRA2/Cor.agda) with a primitive
recursive defining equation; for value-shaped inputs $x$ it satisfies
$\mathit{ap}_1\;\mathit{cor}\;x \;=\; \mathit{code}\;x$
internally (the lemma $\mathit{corOfReify}$).  At free variables it
does \emph{not} reduce, just as Paulson's $\mathrm{QR}(x, x')$ does
not equate $x$ with anything specific: instead it constrains $x'$
relationally.  Theorem~14 step~3 in our development then proves
internal-implication statements
\[
  \mathit{thmT}\,x = \mathit{codeFormula}\,G \;\;\to\;\;
  \mathit{thmT}\,(g\_\mathit{part}\,x) = \widehat{\mathit{th}\,\hat x = \mathit{sub}\,j\,j},
\]
parametric in $x$, with the $\hat x = \mathit{cor}\,x$ slot left
syntactic and the closure transporting it to $\mathit{code}\,x$ only
at the value-shaped instantiation step (the encode-witness, certified
by $\mathit{IsValue}$).  This is structurally the same move as
Paulson's: introduce parametric new variables, constrain by relation
on the antecedent side, never assert ``$\hat x$ equals the numeral of
$x$'' as an unconditional formal claim.

A further connection to Paulson's \S7: \'{S}wierczkowski needed an
ordering on the HF universe to define $Q$ by recursion on rank, and
the resulting construction had a gap.  In Guard's BRA, $\mathit{cor}$
requires no such ordering --- it is built directly as $\mathit{Rec}\;
\mathit{stepCor}$ from BRA's primitive-recursion combinator on binary
trees (BRA2/Cor.agda), with $\mathit{stepCor}$ a compact $\mathit{Fan}
+ \mathit{Lift} + \mathit{KT} + \mathit{Post} + \mathit{Pair}$
expression.  The primitive-recursive structure of trees is enough;
no recursion on rank, no ordering, no gap.

The present development follows Guard~\cite{guard1963}: the
``provability predicate'' $\mathit{thmT}$ is itself a
primitive-recursive combinator built from $\mathit{Rec} + \mathit{Fan}
+ \mathit{Lift} + \mathit{IfLf}$; encoded substitution is performed by
the explicit primitive recursive functor $\mathit{sub} :
\mathit{Fun}_2$; the entire chain stays first-order; and the proof is
equational at the encoded layer rather than provability-logical at
the formula layer (Section~\ref{sec:internal}).

\begin{thebibliography}{9}

\bibitem{guard1963}
J.~R.~Guard,
\newblock \emph{Lecture notes on recursive arithmetic},
\newblock Applied Mathematics Division, Argonne National Laboratory, 1963.

\bibitem{paulson2014}
L.~C.~Paulson,
\newblock \emph{A machine-assisted proof of G\"odel's incompleteness theorems for the theory of hereditarily finite sets},
\newblock Review of Symbolic Logic, vol.~7 no.~3, 2014, pp.~484--498.

\bibitem{swierczkowski2003}
S.~\'{S}wierczkowski,
\newblock \emph{Finite sets and G\"odel's incompleteness theorems},
\newblock Dissertationes Mathematicae, vol.~422, Polish Academy of Sciences, 2003.

\bibitem{boolos1993}
G.~S.~Boolos,
\newblock \emph{The Logic of Provability},
\newblock Cambridge University Press, 1993.

\bibitem{urbankaliszyk2012}
C.~Urban and C.~Kaliszyk,
\newblock \emph{General bindings and alpha-equivalence in Nominal Isabelle},
\newblock Logical Methods in Computer Science, vol.~8 no.~2, 2012.

\bibitem{daviespfenning2001}
R.~Davies and F.~Pfenning,
\newblock \emph{A modal analysis of staged computation},
\newblock Journal of the ACM, vol.~48 no.~3, May 2001, pp.~555--604.

\bibitem{leivant1999ramified}
D.~Leivant,
\newblock \emph{Ramified recurrence and computational complexity~III: Higher type recurrence and elementary complexity},
\newblock Annals of Pure and Applied Logic, vol.~96 no.~1--3, 1999, pp.~209--229.

\bibitem{wilkieparis1987}
A.~J.~Wilkie and J.~B.~Paris,
\newblock \emph{On the scheme of induction for bounded arithmetic formulas},
\newblock Annals of Pure and Applied Logic, vol.~35, 1987, pp.~261--302.
\newblock See also S.~R.~Buss, \emph{Bounded Arithmetic}, Bibliopolis, 1986.

\end{thebibliography}

\end{document}
