\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{remark}{Remark}

\newcommand{\Term}{\mathrm{Term}}
\newcommand{\Formula}{\mathrm{Formula}}
\newcommand{\Deriv}{\mathsf{Deriv}}
\newcommand{\thmT}{\mathsf{thmT}}
\newcommand{\evalU}{\mathsf{evalU}}
\newcommand{\parse}{\mathsf{parse}}
\newcommand{\Kcode}{\mathsf{Kcode}}
\newcommand{\outK}{\mathsf{outKdef}}
\newcommand{\closeW}{\mathsf{closeW}}
\newcommand{\codeFalse}{\mathsf{codeFalse}}
\newcommand{\codeF}{\mathsf{code}}
\newcommand{\num}{\mathsf{num}}
\newcommand{\ap}[1]{\mathsf{ap}_{#1}}
\newcommand{\Fun}[1]{\mathsf{Fun}_{#1}}
\newcommand{\cgFun}{\mathsf{cgFun}}
\newcommand{\imp}{\mathbin{\Rightarrow}}
\newcommand{\eqF}{\mathsf{eqF}}
\newcommand{\GG}{\mathsf{G}}
\newcommand{\Rf}{R}
\newcommand{\Lstar}{L^{\!\star}}
\newcommand{\kmax}{k_{\max}}
\newcommand{\ruleInst}{\mathsf{ruleInst}}
\newcommand{\compile}{\mathsf{compile}}
\newcommand{\denote}{\mathsf{denote}}
\newcommand{\Exp}{\mathsf{Exp}}
\newcommand{\gFact}{\mathsf{gFact}}

\title{Chaitin--G\"odel I in BRA4:\\
  a genuine $\Fun{1}$ diagonal with an internal implication}
\author{}
\date{}

\begin{document}
\maketitle

\section{Chaitin's argument, recalled}

Chaitin's proof of G\"odel's first incompleteness theorem~\cite{Chaitin1971,Chaitin1974}
runs as follows. Fix a sound theory $T$ that interprets enough arithmetic, and fix a
universal partial computable interpreter $U$ with a Kolmogorov-complexity bound
$K_U(x) := \min\{|p| : U(p)\downarrow = x\}$. There is a constant $c$
(depending only on $T$ and $U$) such that, for every $\Lstar$ and every
integer $x'$,
\[
  T \vdash K_U(\underline{x'}) > \Lstar
  \quad\Longrightarrow\quad
  \text{$T$ is inconsistent, provided } \Lstar > c + \log \Lstar.
\]
The contradiction is exhibited by a single short program $g_{\Lstar}$:
enumerate $\thmT(0), \thmT(1), \dots$ until a proof of the form
$\codeF{K_U(\underline{?}) > \Lstar}$ is found, then output the
witnessed integer~$x'$. The size of $g_{\Lstar}$ is $c + \log \Lstar$,
strictly below $\Lstar$, so the very witness $x'$ that $T$ proves to be
$K_U$-incompressible is itself $g_{\Lstar}$-output --- a contradiction
to the size lemma internal to $T$.

\section{The BRA4 statement}

The formalisation packages Chaitin's argument as a single closed
$T$-derivation: it produces, from any putative $T$-proof that
$\thmT\,w$ asserts incompressibility of its own read-back, a $T$-proof
that a constructed code is a proof of $0=1$. Two design choices make
this statement sharp.

\begin{enumerate}
\item The ``construct the short program $g_{\Lstar}$'' step is a
  genuine BRA combinator $\GG : \Fun{1}$, an honest object-level
  program, not a meta-level term transformer. The constructed proof
  code is literally $\ap{1}\,\GG\,w$.
\item The ``if $\dots$ then inconsistent'' step is an
  \emph{internal} Hilbert implication $\imp$ inside a single $\Deriv$,
  not a meta-level function $\Deriv\,X \to \Deriv\,Y$. The deduction
  content lives inside the object logic.
\end{enumerate}

The headline theorem
(\texttt{BRA4/CgFunFun1.agda}, \texttt{chaitinG1\_Fun1\_all}) is:
\begin{align*}
  \textsf{chaitinG1\_Fun1\_all} :{}
    & (w : \Term) \to {} \\
    & \Deriv\,\bigl(\,
        \eqF\,(\ap{1}\,\thmT\,w)\,
              (\ap{1}\,(\Kcode_{\Lstar})\,(\ap{1}\,\outK_{\Lstar}\,w))
        \;\imp\;
        \eqF\,(\ap{1}\,\thmT\,(\ap{1}\,\GG\,w))\,\codeFalse
      \,\bigr).
\end{align*}
Here $w$ is an \emph{arbitrary} Term: no closure hypothesis, no
\textsf{isNat} integrality predicate, no $\mathrm{Con}(T)$ premise, and
no closedness bookkeeping. In words: \emph{for every} $w$, $T$ proves
``if $\thmT\,w$ is the incompressibility statement
$\Kcode_{\Lstar}$ about its own read-back $\outK_{\Lstar}\,w$, then the
code $\ap{1}\,\GG\,w$ is a $T$-proof of $0=1$''.

The antecedent $\eqF\,(\ap{1}\,\thmT\,w)\,(\Kcode_{\Lstar}\,(\outK_{\Lstar}\,w))$
is the firing condition in $\num$-raw form: the subject slot is
$\num\,x'$ for an arbitrary Term $x'$, with no integrality required to
align with the read-back. Whether $\thmT\,w$ has the form
$\Kcode_{\Lstar}\,(\_)$ is read off internally by the BRA function
$\outK_{\Lstar} = \mathsf{decode}\circ\mathsf{projKdef}\circ\thmT$, so
the witnessed integer is forced to be $\outK_{\Lstar}\,w$.

\section{The diagonal program \texorpdfstring{$\GG$}{G}}

\subsection{Bracket abstraction: combinatory completeness}

The short program is constructed by a one-variable bracket-abstraction
DSL (\texttt{BRA4/AbsFun1.agda}). An expression $e : \Exp$ is a
Term-valued expression in one distinguished variable, built from
\begin{itemize}
\item the variable leaf $\mathsf{evar}$;
\item constant leaves $\mathsf{econst}\,c$ with $c : \Term$ var-free;
\item object application nodes $\mathsf{eap}_1\,f$ ($f : \Fun{1}$) and
  $\mathsf{eap}_2\,g$ ($g : \Fun{2}$).
\end{itemize}
Two maps interpret $\Exp$:
\[
  \denote : \Exp \to \Term \to \Term
  \qquad\text{(plug $x$ for $\mathsf{evar}$),}
  \qquad
  \compile : \Exp \to \Fun{1}
  \qquad\text{(bracket-abstract $\mathsf{evar}$),}
\]
with $\compile\,\mathsf{evar} = u$, $\compile\,(\mathsf{eap}_1\,f\,e) =
\mathsf{compose1U}\,f\,(\compile\,e)$, and $\compile\,(\mathsf{eap}_2\,g\,a\,b)
= C\,g\,(\compile\,a)\,(\compile\,b)$. The combinatory-completeness fact is a
\emph{proved} $\Deriv$:
\[
  \mathsf{compile\_eq} : (e : \Exp)\,(x : \Term) \to
    \Deriv\,\bigl(\eqF\,(\ap{1}\,(\compile\,e)\,x)\,(\denote\,e\,x)\bigr),
\]
by induction on $e$ from the axioms $\mathsf{ax\_u}$ (variable),
$\mathsf{constTermFun1\_eq}$ (constant), and
$\mathsf{axComp}/\mathsf{axComp2}$ with $\mathsf{cong}$ congruences
(application nodes).

\begin{remark}
The factorisation is genuinely a $\Deriv$, not a meta-level (Agda)
equality. $\ap{1}\,(\compile\,e)\,x$ has head constructor $\ap{1}$,
while $\denote\,e\,x$ generally has a different head (e.g.\ $\ap{2}\,
\mathsf{Pair}$): they are distinct Term normal forms. The combinator
semantics lives only in $\Deriv$, through the $\mathsf{ax\_C}/
\mathsf{axComp}/\mathsf{ax\_u}$ axioms. There is no definitional
collapse to exploit.
\end{remark}

\subsection{The body expression}

The Term that Chaitin's construction produces from $w$ is a fixed
object-level combinator context $E[\,\cdot\,]$ with a single
$w$-dependent slot. Every node of $E$ is an $\ap{1}/\ap{2}$ of an
object $\Fun{1}/\Fun{2}$
($\mathsf{gRec}$, $\mathsf{sigma}$, $\mathsf{pi}$, $\num$,
$\mathsf{Pair}$, the fuel combinators $\mathsf{fuelMu}$, $\mathsf{fuelG}$,
the read-back $\outK_{\Lstar}$, $\mathsf{runProg}$, $s$) or a
var-free constant; the proof-assembly builders
($\mathsf{cmp}$, $\mathsf{cImp}$, $\mathsf{cNeg}$, $\mathsf{cEqTm}$,
$\mathsf{cAp1f}$, $\mathsf{cAp2f}$, $\mathsf{exfProof}$ and its
$\mathsf{axK}/\mathsf{axNeg}/\mathsf{axS}/\mathsf{liftP}/\mathsf{bComb}$
chains) are themselves pure $\ap{2}\,\mathsf{Pair}$-skeletons of their
arguments. Hence $E[\,\cdot\,]$ is a one-variable $\Exp$, transcribed
node-for-node into \texttt{bodyExp} (\texttt{BRA4/CgFunFun1.agda}) using
$\Exp$ smart constructors that mirror each builder
($\mathsf{ecmpE}$, $\mathsf{eexfProofE}$, \dots). The diagonal program
and its defining fact are
\[
  \GG := \compile\,\texttt{bodyExp} : \Fun{1},
  \qquad
  \gFact\,w := \mathsf{compile\_eq}\,\texttt{bodyExp}\,(\closeW\,w).
\]
Here $\closeW\,w := \mathsf{substT}\,0\,O\,(\mathsf{substT}\,1\,O\,w)$
forces the variables $0$ and $1$ in $w$ to $O$; on the denotation side
$\denote\,\texttt{bodyExp}\,(\closeW\,w)$ is the Chaitin Term verbatim,
because the private constants in \texttt{bodyExp} are replicated
exactly. Thus
\[
  \gFact : (w : \Term) \to
    \Deriv\,\bigl(\eqF\,(\ap{1}\,\GG\,(\closeW\,w))\,(\denote\,\texttt{bodyExp}\,(\closeW\,w))\bigr),
\]
a proved object equality identifying the program image $\ap{1}\,\GG$
with the Chaitin construction.

\section{The internal implication \texorpdfstring{$\imp$}{=>}}

The ``if firing condition then inconsistent'' step is an object-level
implication, obtained by internalising the deduction theorem
(\texttt{BRA4/CgFalseImp.agda}). The parametric form carries an
arbitrary antecedent context $\Rf : \Formula$:
\begin{align*}
  \mathsf{cgFalseImp} :{}
    & (\Rf : \Formula)\,(w : \Term) \to {} \\
    & \text{(closedness of $\Rf$ at vars $0,1$)} \to {} \\
    & \Deriv\,\bigl(\Rf \imp
        \eqF\,(\ap{1}\,\thmT\,(\closeW\,w))\,(\Kcode_{\Lstar}\,(\outK_{\Lstar}\,(\closeW\,w)))\bigr) \to {} \\
    & \Deriv\,\bigl(\Rf \imp
        \eqF\,(\ap{1}\,\thmT\,(\cgFun\,w))\,\codeFalse\bigr).
\end{align*}
The implication is a single $\Deriv$ over $\imp$; no meta-level arrow
$\Deriv\,X \to \Deriv\,Y$ ever appears. The antecedent context $\Rf$
threads through the $\mathsf{ruleIndNat}$ motive $Q := \Rf \imp P$ used
in the operational core, and through the substitution bridges, with the
closedness hypotheses on $\Rf$ supplying what those steps need.

Setting $\Rf$ to the firing condition itself and applying internal
reflexivity gives the un-parametrised form
$\mathsf{cgFalseImpDed}$, and instantiating at the fresh witness
$w := \mathsf{var}\,2$ gives the closed Hilbert formula
$\mathsf{cgFalseImpAtVar2}$:
\[
  \Deriv\,\bigl(
    \eqF\,(\ap{1}\,\thmT\,v_2)\,(\Kcode_{\Lstar}\,(\outK_{\Lstar}\,v_2))
    \;\imp\;
    \eqF\,(\ap{1}\,\thmT\,(\cgFun\,v_2))\,\codeFalse
  \bigr),
  \qquad v_2 = \mathsf{var}\,2.
\]
At $v_2$ all three substitution-closedness witnesses are reflexivity
($\mathsf{substT}$ on $\mathsf{var}\,2$ leaves it fixed since $0,1 \neq
2$), so $\closeW\,v_2 = v_2$ definitionally.

\section{Assembling the headline}

The headline $\textsf{chaitinG1\_Fun1\_all}$ combines the two pieces at
the fresh witness and then instantiates. The construction:
\begin{enumerate}
\item At $v_2 = \mathsf{var}\,2$ we have $\closeW\,v_2 = v_2$, hence
  $\gFact\,v_2 : \Deriv\,(\eqF\,(\ap{1}\,\GG\,v_2)\,(\cgFun\,v_2))$ ---
  the program image \emph{is} the Chaitin construction at $v_2$.
\item Re-index the conclusion's subject from $\cgFun\,v_2$ to
  $\ap{1}\,\GG\,v_2$ under $\thmT$: from
  $\mathsf{cong}_1\,\thmT\,(\mathsf{ruleSym}\,(\gFact\,v_2))$ and
  $\mathsf{ax\_eqTrans}$ one builds
  $\Deriv\,(\eqF\,(\thmT\,(\cgFun\,v_2))\,\codeFalse \imp
            \eqF\,(\thmT\,(\ap{1}\,\GG\,v_2))\,\codeFalse)$, and
  composes it after $\mathsf{cgFalseImpAtVar2}$ (via $\mathsf{compI}$),
  yielding the implication at $v_2$ with subject $\ap{1}\,\GG\,v_2$.
\item Apply $\ruleInst\,(2 := w)$. Because $v_2$ is the sole occurrence
  and $\ap{1}\,\GG$ is mere data,
  \[
    \mathsf{substT}\,2\,w\,(\ap{1}\,\GG\,v_2)
      = \ap{1}\,\GG\,(\mathsf{substT}\,2\,w\,v_2)
      = \ap{1}\,\GG\,w,
  \]
  and the antecedent specialises to the firing condition at $w$. The
  result holds for \emph{every} $w$.
\end{enumerate}

\begin{remark}[Why no closedness is needed]
A genuine $\Fun{1}$ program is what removes the closedness bookkeeping.
The substitution $\ruleInst\,(2 := w)$ commutes cleanly through the
\emph{data} node $\ap{1}\,\GG$, so the subject becomes exactly
$\ap{1}\,\GG\,w$ for arbitrary $w$. A meta-level term transformer would
embed $\closeW\,w$ in its internal slots, and substituting
$v_2 \mapsto w$ into the transformer's value at $v_2$ would not
reproduce its value at $w$; that mismatch is precisely what forced
closedness hypotheses in any term-transformer formulation. With an
honest combinator $\GG$, the firing point is a fresh variable and
the conclusion follows for all $w$ by one universal instantiation.
\end{remark}

\section{The subtlety: the internal evaluator and majorisation}

The conclusion ultimately rests on an \emph{object-level} run of the
universal interpreter $\evalU$ on the looping program $g_{\Lstar}$
embedded in $\GG$. This is where the formalisation is delicate.

\subsection{Representation of the universal interpreter}

Inside BRA4, $\evalU$ is realised as a small CK-machine:
configurations carry a control Term together with a continuation
stack, and a one-step transition function $\mathsf{stepU}$ advances one
rule at a time. Running the interpreter for fuel $n$ is iteration of
$\mathsf{stepU}$ from a starting configuration. A second layer wraps
the $\mu$-search loop of $g_{\Lstar}$, so that the interpreter's outer
behaviour matches Chaitin's informal description ``enumerate $\thmT\,0,
\thmT\,1, \dots$ until a fitting proof is found''. The rule-by-rule
layout of $\mathsf{stepU}$ is not what makes the proof delicate; what
matters is the \emph{shape} of the completeness statement the proof
needs.

\subsection{Gandy/Howard majorisation for Term-fuel completeness}

A first attempt at a completeness statement would say: for every
\emph{numeral} fuel $\overline{n}$ large enough, iterating
$\mathsf{stepU}$ on $g_{\Lstar}$ at $\overline{n}$ reaches the final
configuration $\mathsf{cfgRT}\,(\num\,x')\,K_0$. Such a statement is
useless inside an internal $\Deriv$-argument: $\thmT$, $\evalU$, and
$\mathsf{stepU}$ all take Terms, not meta-Nats, so the fuel slot of the
resulting $\thmT$-fact must itself be a Term. The completeness theorem
actually needed has the form
\[
  (x : \Term) \to \Deriv\bigl(
    \mathsf{iter}\,\mathsf{stepU}\,
      (\mathsf{cfgEV}\,(\mathsf{mcode}\,f)\,x\,K_0)\,
      (\mathsf{fuelG}\,f\,x)
      = \mathsf{cfgRT}\,(\num\,(f\,x))\,K_0
  \bigr),
\]
i.e.\ Term-input/Term-fuel completeness, where $\mathsf{fuelG}\,f$ is a
closed BRA combinator. Proving such a statement by structural induction
on the combinator tree of $f$ stalls at the $R$-recursion node: the
inner induction would need numerical control over how much fuel the
recursive calls consume, but the actual running time is not available
at the $\Deriv$-level.

The resolution is the standard Gandy/Howard majorisation argument. One
builds, by primitive recursion on the structure of $f$, a closed BRA
combinator $\mathsf{fuelG}\,f : \Fun{1}$ that provably
\emph{majorises} the numerical fuel any actual $\mathsf{stepU}$-run
consumes at every Term input. The induction-with-substituted-motive
then carries the stable shape ``the run completes at \emph{any} fuel
$\geq \mathsf{fuelG}\,f\,x$'', which descends through the $R$-node
because $\mathsf{fuelG}$ is itself defined by a recursive scheme that
mirrors $f$. The $\mu$-loop wrapping $g_{\Lstar}$ requires the same
idea one level higher: a majorising combinator $\mathsf{fuelMu}$
controls the outer search depth and is layered on top of
$\mathsf{fuelG}$ by the fuel-composition lemma $\mathsf{iter\_add\_T}$.
The closed Term fuel that finally appears inside $\ap{1}\,\GG\,w$ is
\[
  n_T \;:=\; \mathsf{fuelMu}\,\kmax\,\kmax \;+\; \mathsf{fuelG}\,\kmax\,O,
  \qquad
  \kmax = \mathsf{gRec}\,O\,(s\,(\closeW\,w)),
\]
and the resulting positive $\thmT$-fact
\[
  \thmT\,\mathsf{cPos} \stackrel{\Deriv}{=}
    \codeF{\mathsf{runProg}\,g_L\,n_T = s\,x'}
\]
follows from this run via Theorem~13 (singulary). Without majorisation
there would simply be no Term to put in the fuel slot.

\subsection{Closing the chain}

With Term-fuel completeness in hand, the operational core closes the
chain
\[
  \evalU\,(\parse\,\mathsf{gLnameDef})\,n_T = s\,(\outK_{\Lstar}\,\kmax)
\]
through seven internal segments (the module $\mathsf{ChainKdef}$),
glued by $\mathsf{iter\_add\_T}$. The positive $\thmT$-fact above is its
internalisation via Theorem~13. The negative leg
$\thmT\,(\mathsf{cmp}\,\mathsf{outerWrap}\,\mathsf{cSize}) =
\mathsf{cNeg}\,\codeF{\mathsf{runProg}\,g_L\,n_T = s\,x'}$ is obtained
from the firing condition by two $\mathsf{thmT\_at\_sb}$ substitutions
($\mathsf{passK}$) and an $\mathsf{encoded\_mp}$ against the size atom.
The clash is then a single
$\mathsf{encoded\_mp}\,/\,\mathsf{encoded\_exfalso}$ pair, internal
throughout under the $\imp$-context.

\subsection{No consistency hypothesis}

Chaitin's original argument concludes ``$T$ is inconsistent'' (in the
metatheory). The BRA4 formalisation strengthens this: the conclusion is
an \emph{internal} implication
$\Deriv\,(X(w) \imp \eqF\,(\thmT\,(\ap{1}\,\GG\,w))\,\codeFalse)$,
i.e.\ a single BRA derivation that, on the firing condition, the
constructed code $\ap{1}\,\GG\,w$ is a $T$-proof of $0=1$. No
$\mathrm{Con}(T)$ premise is consumed, no \textsf{isNat} integrality
predicate is consumed, and the firing condition is stated in $\num$-raw
form.

This is the precise interface offered to the later surprise-G\"odel-II
descent: that descent will \emph{not} need to reprove any evaluator
completeness, and will use the implication purely as a black-box arrow
from a $\Deriv$-witness of the firing condition to a $\Deriv$-witness of
$\codeFalse$. The Term-fuel completeness work paid for here --- and in
particular the Gandy/Howard machinery of \S6.2 --- is invisible to the
descent. The contrast with consistency is sharp: the implication itself
consumes no $\mathrm{Con}(T)$, but the surprise-G\"odel-II argument will
consume internal consistency \emph{after}, and only after, the
implication has been discharged.

\section{Files}

\begin{itemize}
\item \texttt{BRA4/AbsFun1.agda} --- the one-variable bracket-abstraction
  DSL: $\Exp$, $\denote$, $\compile$, and the combinatory-completeness
  evaluation $\mathsf{compile\_eq}$ (a proved $\Deriv$).
\item \texttt{BRA4/CgFunFun1.agda} --- the genuine $\Fun{1}$ diagonal
  $\GG = \compile\,\texttt{bodyExp}$, its defining fact $\gFact$, and
  the headline $\textsf{chaitinG1\_Fun1\_all}$ (plus the $\Rf$-carrying
  and the closed-witness corollaries).
\item \texttt{BRA4/CgFalseImp.agda} --- the internal implication
  $\mathsf{cgFalseImp}$ (deduction-theorem internalisation), its
  un-parametrised form $\mathsf{cgFalseImpDed}$, and the fresh-witness
  instance $\mathsf{cgFalseImpAtVar2}$.
\item \texttt{BRA4/CloseW.agda} --- $\closeW$ and the structural lemmas
  it supports.
\end{itemize}

All files typecheck under \texttt{--safe --without-K --exact-split}
with no holes and no postulates.

\begin{thebibliography}{9}

\bibitem{Chaitin1971}
G.~J. Chaitin, ``Computational complexity and G\"odel's incompleteness theorem,''
\emph{ACM SIGACT News}, no.~9, pp.~11--12, 1971.

\bibitem{Chaitin1974}
G.~J. Chaitin, ``Information-theoretic limitations of formal systems,''
\emph{Journal of the ACM}, vol.~21, no.~3, pp.~403--424, 1974.

\end{thebibliography}

\end{document}
